---
title: "Distributed Lag Non-Linear Model (DLNM) Analysis"
subtitle: "Daily Precipitation and Asthma ED Visits in NYC (2016–2019)"
author: "Minseo (Brenda) Kim"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

Load libraries

```{r}
library(dlnm)
library(splines)
library(lubridate)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
```

# Data Preparation

## Load Data

Borough-level asthma ED visits, precipitation, and PM2.5 data (2016–2019). Daily mean temperature from NOAA GHCN, Central Park station (USW00094728).

```{r}
borough_data <- read_csv("datasets/asthma_precip_pm25_clean.csv")
temp_data    <- read_csv("datasets/daily_temp_nyc.csv")

glimpse(borough_data)
summary(borough_data)

glimpse(temp_data)
```

## Aggregate to City-Wide

DLNM is fit at the city-wide level. Summing asthma counts and averaging exposure across boroughs per day.

```{r}
citywide <- borough_data %>%
  mutate(date = as.Date(date)) %>%
  group_by(date) %>%
  summarise(
    asthma_total = sum(asthma_er_count, na.rm = TRUE),
    ppt_mean     = mean(ppt, na.rm = TRUE),
    pm25_mean    = mean(pm25, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(date)
```

## Merge Temperature

Central Park station serves as the city-wide temperature reference (single-station design typical of NYC time-series studies — temperature gradients across the city are small relative to the daily variation we model). Mean temperature is computed as (TMAX + TMIN) / 2 in °F.

```{r}
temp_clean <- temp_data %>%
  mutate(
    date      = as.Date(DATE),
    tmax      = as.numeric(TMAX),
    tmin      = as.numeric(TMIN),
    temp_mean = (tmax + tmin) / 2
  ) %>%
  select(date, tmax, tmin, temp_mean)

citywide <- citywide %>%
  left_join(temp_clean, by = "date")

# Date range and completeness check
cat("Date range:", as.character(min(citywide$date)), "to", as.character(max(citywide$date)), "\n")
cat("Total days:", nrow(citywide), "\n")
cat("Expected days:", as.numeric(max(citywide$date) - min(citywide$date)) + 1, "\n")

full_dates <- seq.Date(min(citywide$date), max(citywide$date), by = "day")
missing_dates <- full_dates[!full_dates %in% citywide$date]
cat("Missing dates:", length(missing_dates), "\n")

cat("Missing temperature (mean):", sum(is.na(citywide$temp_mean)), "\n")
cat("Missing precipitation:",      sum(is.na(citywide$ppt_mean)),  "\n")
cat("Missing PM2.5:",               sum(is.na(citywide$pm25_mean)), "\n")
```

## Create Time Variables

Day of week, season, and numeric time index for the long-term trend spline.

```{r}
citywide <- citywide %>%
  mutate(
    year        = year(date),
    month       = month(date),
    dow         = factor(wday(date, label = TRUE)),
    season      = factor(case_when(
      month %in% c(12, 1, 2)  ~ "Winter",
      month %in% c(3, 4, 5)   ~ "Spring",
      month %in% c(6, 7, 8)   ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    ), levels = c("Winter", "Spring", "Summer", "Fall")),
    time        = as.numeric(date - min(date))
  )
```

# Exploratory Data Analysis

## Summary Statistics

```{r}
citywide %>%
  summarise(
    across(
      c(asthma_total, ppt_mean, pm25_mean, temp_mean, tmax, tmin),
      list(
        mean   = ~mean(., na.rm = TRUE),
        sd     = ~sd(., na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        min    = ~min(., na.rm = TRUE),
        max    = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  print(n = Inf)
```

## Precipitation Distribution

Key percentiles and histogram with 95th percentile marked.

```{r}
precip_quantiles <- quantile(citywide$ppt_mean,
                              probs = c(0.25, 0.50, 0.75, 0.90, 0.95, 0.99),
                              na.rm = TRUE)
cat("Precipitation percentiles (mm):\n")
print(round(precip_quantiles, 2))

cat("\nZero-precipitation days:",
    round(mean(citywide$ppt_mean == 0, na.rm = TRUE) * 100, 1), "%\n")

ggplot(citywide, aes(x = ppt_mean)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = precip_quantiles["95%"], linetype = "dashed", color = "red") +
  annotate("text", x = precip_quantiles["95%"], y = Inf, label = "95th percentile",
           vjust = 2, hjust = -0.1, color = "red", size = 3.5) +
  labs(title = "City-Wide Daily Mean Precipitation",
       x = "Precipitation (mm)", y = "Frequency") +
  theme_minimal()
```

The precipitation distribution is extremely right-skewed with a large spike at 0 mm (median 0; \~45.7% of days), a modest amount of light precipitation (75th percentile 2.69 mm), and a sparse extreme tail; the dashed red line marks the 95th percentile (20.56 mm), highlighting that estimates for very heavy rain (\>20-80mm) are based on sparse data in the tail.

## Cross-Correlation

CCF between precipitation and asthma ED visits (up to 21-day lag).

```{r}
ccf_result <- ccf(citywide$ppt_mean, citywide$asthma_total,
                   lag.max = 21, plot = FALSE)

ccf_df <- data.frame(lag = ccf_result$lag, acf = ccf_result$acf)

ggplot(ccf_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_hline(yintercept = c(-1, 1) * qnorm(0.975) / sqrt(nrow(citywide)),
             linetype = "dashed", color = "blue") +
  geom_segment(aes(xend = lag, yend = 0), color = "steelblue") +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "CCF: Precipitation vs. Asthma ED Visits",
       subtitle = "Positive lag = precipitation leads",
       x = "Lag (days)", y = "Cross-Correlation") +
  theme_minimal()
```

Cross-correlations are near zero at most lags (within the plotted significance bounds), implying weak linear lead–lag association in raw series. There is one visibly stronger negative correlation appears around a short positive lag (precipitation leading by \~1 day), but its magnitude remains small in absolute terms. =\> No strong linear lead-lag relationship between daily rain and asthma ED visits in the raw data.

## Autocorrelation of Asthma ED Visits

```{r}
acf(citywide$asthma_total, lag.max = 21,
    main = "ACF: Daily Asthma ED Visits (City-Wide)")
```

Autocorrelation is strongly positive across many lags with a slow decay and apparent weekly structure (elevations around \~7, \~14, \~21 days). It confirms substantial temporal dependence (day-of-week and longer-term patterns), which the DLNM model correctly controls for using the time spline, DOW, and season terms.

## Temperature Distribution

Daily mean temperature (°F) at Central Park.

```{r}
temp_quantiles <- quantile(citywide$temp_mean,
                            probs = c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99),
                            na.rm = TRUE)
cat("Temperature percentiles (°F):\n")
print(round(temp_quantiles, 1))

ggplot(citywide, aes(x = temp_mean)) +
  geom_histogram(bins = 50, fill = "darkorange", color = "white", alpha = 0.8) +
  geom_vline(xintercept = temp_quantiles[c("1%", "99%")],
             linetype = "dashed", color = "red") +
  annotate("text", x = temp_quantiles["1%"],  y = Inf, label = "1st pctl",
           vjust = 2, hjust = -0.1, color = "red", size = 3.5) +
  annotate("text", x = temp_quantiles["99%"], y = Inf, label = "99th pctl",
           vjust = 2, hjust = 1.1,  color = "red", size = 3.5) +
  labs(title = "City-Wide Daily Mean Temperature (Central Park)",
       x = "Mean Temperature (°F)", y = "Frequency") +
  theme_minimal()
```

Temperature shows the expected bimodal-ish seasonal distribution centered around ~55°F, with cold-season days and hot-season days defining the extreme tails. These percentiles will anchor the lag-response slices in the temperature DLNM.

## Precipitation–Temperature Relationship

Quick check on whether precipitation and temperature are correlated enough to warrant joint modeling.

```{r}
cor_pt <- cor(citywide$ppt_mean, citywide$temp_mean,
              use = "complete.obs", method = "spearman")
cat("Spearman correlation (precip, temp):", round(cor_pt, 3), "\n")

ggplot(citywide, aes(x = temp_mean, y = ppt_mean)) +
  geom_point(alpha = 0.3, color = "steelblue", size = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "darkred", span = 0.5) +
  labs(title = "Daily Precipitation vs. Mean Temperature",
       x = "Mean Temperature (°F)", y = "Precipitation (mm)") +
  theme_minimal()
```

Precipitation and temperature are weakly correlated in raw daily values, but they share strong seasonal structure. This justifies including temperature as a separate adjustment in the DLNM rather than relying solely on the season term and time spline to absorb its effect.

# DLNM Model

## Cross-Bases

Parallel natural cubic spline crossbases for precipitation and temperature: 3 df for the exposure-response dimension, 4 df for the lag-response dimension, max lag = 14 days.

```{r}
cb_precip <- crossbasis(
  citywide$ppt_mean,
  lag    = 14,
  argvar = list(fun = "ns", df = 3),
  arglag = list(fun = "ns", df = 4)
)

cb_temp <- crossbasis(
  citywide$temp_mean,
  lag    = 14,
  argvar = list(fun = "ns", df = 3),
  arglag = list(fun = "ns", df = 4)
)

summary(cb_precip)
summary(cb_temp)
```

## Primary Model

Quasi-Poisson GLM: `log(E[Y]) = α + cb(precip, lag 0–14) + cb(temp, lag 0–14) + ns(time, 7df/yr) + DOW + season`. Temperature enters as a confounder of the precipitation–asthma association (it drives both rainfall meteorology and asthma incidence); PM2.5 is excluded from the primary model as a hypothesized mediator (per DAG-based advisor feedback).

```{r}
n_years <- as.numeric(max(citywide$date) - min(citywide$date)) / 365.25
time_df <- round(7 * n_years)

model_primary <- glm(
  asthma_total ~ cb_precip + cb_temp +
    ns(time, df = time_df) +
    dow +
    season,
  family = quasipoisson(link = "log"),
  data   = citywide
)

summary(model_primary)
cat("\nDispersion parameter:", summary(model_primary)$dispersion, "\n")
```

## DLNM Predictions

Precipitation centered at 0 mm (no rain). Temperature centered at the median (~55 °F), the natural reference for a bidirectional exposure where both cold and heat may elevate risk.

```{r}
pred_primary <- crosspred(
  cb_precip,
  model_primary,
  at = seq(0, max(citywide$ppt_mean, na.rm = TRUE), length.out = 100),
  cen = 0
)

temp_ref <- median(citywide$temp_mean, na.rm = TRUE)
cat("Temperature reference (median):", round(temp_ref, 1), "°F\n")

pred_temp <- crosspred(
  cb_temp,
  model_primary,
  at  = seq(min(citywide$temp_mean, na.rm = TRUE),
            max(citywide$temp_mean, na.rm = TRUE),
            length.out = 100),
  cen = temp_ref
)
```

# Visualizations

## 3D Exposure-Lag-Response Surface

```{r, fig.height=8}
plot(pred_primary,
     xlab = "Precipitation (mm)", ylab = "Lag (days)", zlab = "RR",
     main = "3D Exposure-Lag-Response:\nPrecipitation and Asthma ED Visits",
     theta = 220, phi = 30, ltheta = -120,
     shade = 0.3, col = "steelblue")
```

The 3D surface shows the full non-linear, lag-dependent relationship between precipitation and asthma ED visit risk (RR) over 0–14 days. Risk stays close to 1.0 across most of the surface. Slight elevations (RR \> 1) appear at higher precipitation levels and short lags (especially lag 1–5 days), while some areas of RR \< 1 are visible at moderate rain and longer lags. The surface is noisy and uncertain at very high precipitation (\>40 mm) because of limited data.

## Contour Plot

```{r}
plot(pred_primary, "contour",
     xlab = "Precipitation (mm)", ylab = "Lag (days)",
     main = "Contour: Precipitation-Lag-Response",
     key.title = title("RR"))
```

The contour plot reinforces the non-linear, lag-varying pattern. Regions of RR \> 1 (higher risk) are concentrated at higher precipitation (\>30–40 mm) and short lags (0–5 days). Regions of RR \< 1 (lower risk) appear at moderate rain levels and longer lags. The effect weakens and sometimes reverses after \~7 days, indicating a complex, short-lived response rather than a simple delayed increase.

## Lag-Specific Exposure-Response

Exposure-response at lag 0, 2, 5, 7, 14 days.

```{r}
lag_days <- c(0, 2, 5, 7, 14)

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
for (lag_val in lag_days) {
  plot(pred_primary, "slices",
       var = NULL, lag = lag_val,
       xlab = "Precipitation (mm)", ylab = "RR",
       main = paste0("Lag ", lag_val, " days"),
       col = "steelblue", ci.arg = list(col = adjustcolor("steelblue", 0.2)))
}
par(mfrow = c(1, 1))
```

The relationship between daily precipitaiton (0-80+mm) and asthma ED visit risk (RR) changes depending on the time lag after the rain. All curves are non-linear, with a general pattern of slight protective effects at low-to-moderate rainfall and increasing risk at heavier rainfall. The strength and timing of the harmful effect vary clearly by lag. Overall, the heavy rainfall (\>40–50 mm) is linked to a modest short-term increase in asthma ED visits that peaks around 2 days later, then gradually weakens and mostly disappears by 7–14 days. Light rain sometimes shows a small protective association, especially at shorter and longer lags. The confidence intervals widen at higher precipitation due to fewer heavy rain days in the data.

## Precipitation-Specific Lag-Response

Lag-response at 50th, 90th, 95th, 99th percentile precipitation values.

```{r}
# Snap quantiles to nearest value in prediction grid
precip_quantile_vals <- quantile(citywide$ppt_mean,
                                  probs = c(0.50, 0.90, 0.95, 0.99),
                                  na.rm = TRUE)
pred_grid <- pred_primary$predvar
precip_vals <- sapply(precip_quantile_vals, function(v) {
  pred_grid[which.min(abs(pred_grid - v))]
})
names(precip_vals) <- names(precip_quantile_vals)

# Reconstruct lag sequence from crosspred object
lag_seq <- seq(pred_primary$lag[1], pred_primary$lag[2], by = pred_primary$bylag)

# Extract and plot each slice manually
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (i in seq_along(precip_vals)) {
  idx <- which.min(abs(pred_primary$predvar - precip_vals[i]))
  rr  <- pred_primary$matRRfit[idx, ]
  lo  <- pred_primary$matRRlow[idx, ]
  hi  <- pred_primary$matRRhigh[idx, ]

  plot(lag_seq, rr,
       type = "l", col = "darkred", lwd = 2,
       xlab = "Lag (days)", ylab = "RR",
       main = paste0("Precip = ", round(precip_vals[i], 1), " mm (",
                      names(precip_vals)[i], ")"),
       ylim = range(c(lo, hi, 1), na.rm = TRUE))
  polygon(c(lag_seq, rev(lag_seq)), c(hi, rev(lo)),
          col = adjustcolor("darkred", 0.2), border = NA)
  lines(lag_seq, rr, col = "darkred", lwd = 2)
  abline(h = 1, lty = 2, col = "gray50")
}
par(mfrow = c(1, 1))
```

How the lag-response pattern (RR over 0–14 days) changes depending on the intensity of precipitation. The reference is 0 mm (no rain). The effect is clearly non-linear and driven mainly by heavier rain events. Overall, typical/moderate rainfall (up to 90th percentile) has essentially no effect on asthma ED visits. Only heavy to very heavy rain (≥95th percentile) produces a detectable short-term rise in risk that peaks around 2 days later and largely disappears within a week. The net cumulative effect over 14 days stays neutral or slightly protective.

## Cumulative Exposure-Response

Total effect of precipitation summed over lag 0–14 days.

```{r}
plot(pred_primary, "overall",
     xlab = "Precipitation (mm)",
     ylab = "Cumulative RR (lag 0–14)",
     main = "Cumulative: Precipitation and Asthma ED Visits",
     col = "steelblue",
     ci.arg = list(col = adjustcolor("steelblue", 0.2)),
     lwd = 2)
abline(h = 1, lty = 2, col = "gray50")
```

Cumulated over the full 14-day lag window, higher precipitation is associated with a modest protective (or null) overall effect. The curve is slightly decreasing up to \~20 mm then flattens, indicating that while short lags show small risk increases, the net 14-day effect is neutral to slightly protective.

## Temperature: Exposure-Lag-Response Surface

```{r, fig.height=8}
plot(pred_temp,
     xlab = "Temperature (°F)", ylab = "Lag (days)", zlab = "RR",
     main = "3D Exposure-Lag-Response:\nTemperature and Asthma ED Visits",
     theta = 220, phi = 30, ltheta = -120,
     shade = 0.3, col = "darkorange")
```

## Temperature: Cumulative Exposure-Response

```{r}
plot(pred_temp, "overall",
     xlab = "Temperature (°F)",
     ylab = "Cumulative RR (lag 0–14)",
     main = "Cumulative: Temperature and Asthma ED Visits",
     col = "darkorange",
     ci.arg = list(col = adjustcolor("darkorange", 0.2)),
     lwd = 2)
abline(h = 1, lty = 2, col = "gray50")
abline(v = temp_ref, lty = 3, col = "gray30")
```

## Temperature: Lag-Response at Cold and Hot Extremes

Lag-response slices at 1st, 50th, and 99th percentile temperature values.

```{r}
temp_quantile_vals <- quantile(citywide$temp_mean,
                                probs = c(0.01, 0.50, 0.99),
                                na.rm = TRUE)
temp_grid  <- pred_temp$predvar
temp_vals  <- sapply(temp_quantile_vals, function(v) {
  temp_grid[which.min(abs(temp_grid - v))]
})
names(temp_vals) <- names(temp_quantile_vals)

lag_seq_t <- seq(pred_temp$lag[1], pred_temp$lag[2], by = pred_temp$bylag)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (i in seq_along(temp_vals)) {
  idx <- which.min(abs(pred_temp$predvar - temp_vals[i]))
  rr  <- pred_temp$matRRfit[idx, ]
  lo  <- pred_temp$matRRlow[idx, ]
  hi  <- pred_temp$matRRhigh[idx, ]

  pct_label <- names(temp_vals)[i]
  panel_color <- if (pct_label == "1%") "dodgerblue" else if (pct_label == "99%") "darkred" else "gray40"

  plot(lag_seq_t, rr,
       type = "l", col = panel_color, lwd = 2,
       xlab = "Lag (days)", ylab = "RR",
       main = paste0("Temp = ", round(temp_vals[i], 1), " °F (", pct_label, ")"),
       ylim = range(c(lo, hi, 1), na.rm = TRUE))
  polygon(c(lag_seq_t, rev(lag_seq_t)), c(hi, rev(lo)),
          col = adjustcolor(panel_color, 0.2), border = NA)
  lines(lag_seq_t, rr, col = panel_color, lwd = 2)
  abline(h = 1, lty = 2, col = "gray50")
}
par(mfrow = c(1, 1))
```

The lag-response slices reveal how cold and hot effects unfold differently over time.

# Sensitivity Analyses

## Varying Max Lag (7, 14, 21 days)

```{r}
lag_windows <- c(7, 14, 21)
models_lag <- list()
preds_lag  <- list()

for (max_lag in lag_windows) {
  cb_temp_p <- crossbasis(
    citywide$ppt_mean,
    lag    = max_lag,
    argvar = list(fun = "ns", df = 3),
    arglag = list(fun = "ns", df = 4)
  )

  cb_temp_t <- crossbasis(
    citywide$temp_mean,
    lag    = max_lag,
    argvar = list(fun = "ns", df = 3),
    arglag = list(fun = "ns", df = 4)
  )

  mod_temp <- glm(
    asthma_total ~ cb_temp_p + cb_temp_t +
      ns(time, df = time_df) +
      dow + season,
    family = quasipoisson(link = "log"),
    data   = citywide
  )

  pred_temp <- crosspred(
    cb_temp_p, mod_temp,
    at  = seq(0, max(citywide$ppt_mean, na.rm = TRUE), length.out = 100),
    cen = 0
  )

  models_lag[[as.character(max_lag)]] <- mod_temp
  preds_lag[[as.character(max_lag)]]  <- pred_temp
}

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (max_lag in lag_windows) {
  plot(preds_lag[[as.character(max_lag)]], "overall",
       xlab = "Precipitation (mm)",
       ylab = paste0("Cumulative RR (lag 0–", max_lag, ")"),
       main = paste0("Max lag = ", max_lag, " days"),
       col = "steelblue",
       ci.arg = list(col = adjustcolor("steelblue", 0.2)),
       lwd = 2)
  abline(h = 1, lty = 2, col = "gray50")
}
par(mfrow = c(1, 1))
```

## Varying Time Spline df (5, 7, 10 df/yr)

```{r}
df_per_year <- c(5, 7, 10)
models_df   <- list()
preds_df    <- list()

for (df_yr in df_per_year) {
  this_df <- round(df_yr * n_years)

  mod_temp <- glm(
    asthma_total ~ cb_precip + cb_temp +
      ns(time, df = this_df) +
      dow + season,
    family = quasipoisson(link = "log"),
    data   = citywide
  )

  pred_temp <- crosspred(
    cb_precip, mod_temp,
    at  = seq(0, max(citywide$ppt_mean, na.rm = TRUE), length.out = 100),
    cen = 0
  )

  models_df[[as.character(df_yr)]] <- mod_temp
  preds_df[[as.character(df_yr)]]  <- pred_temp
}

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (df_yr in df_per_year) {
  plot(preds_df[[as.character(df_yr)]], "overall",
       xlab = "Precipitation (mm)",
       ylab = "Cumulative RR (lag 0–14)",
       main = paste0(df_yr, " df/yr"),
       col = "steelblue",
       ci.arg = list(col = adjustcolor("steelblue", 0.2)),
       lwd = 2)
  abline(h = 1, lty = 2, col = "gray50")
}
par(mfrow = c(1, 1))
```

## Threshold Analysis (95th Percentile Binary)

Binary DLNM: above vs. below 95th percentile precipitation.

```{r}
p90 <- quantile(citywide$ppt_mean, 0.90, na.rm = TRUE)
p95 <- quantile(citywide$ppt_mean, 0.95, na.rm = TRUE)
p99 <- quantile(citywide$ppt_mean, 0.99, na.rm = TRUE)

citywide <- citywide %>%
  mutate(
    precip_above_90 = as.numeric(ppt_mean >= p90),
    precip_above_95 = as.numeric(ppt_mean >= p95),
    precip_above_99 = as.numeric(ppt_mean >= p99)
  )

cb_binary95 <- crossbasis(
  citywide$precip_above_95,
  lag    = 14,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

model_binary95 <- glm(
  asthma_total ~ cb_binary95 + cb_temp +
    ns(time, df = time_df) +
    dow + season,
  family = quasipoisson(link = "log"),
  data   = citywide
)

pred_binary95 <- crosspred(cb_binary95, model_binary95, at = 1, cen = 0)

plot(pred_binary95, "slices", var = 1,
     xlab = "Lag (days)", ylab = "RR",
     main = "Lag-Response: Precip >= 95th Percentile vs. Below",
     col = "darkred",
     ci.arg = list(col = adjustcolor("darkred", 0.2)),
     lwd = 2)
abline(h = 1, lty = 2, col = "gray50")

cat("\nCumulative RR (precip >= 95th pctl, lag 0-14):\n")
cat("  RR:", round(pred_binary95$allRRfit["1"], 3), "\n")
cat("  95% CI:", round(pred_binary95$allRRlow["1"], 3), "-",
    round(pred_binary95$allRRhigh["1"], 3), "\n")
```

## Season-Stratified Models

Separate DLNM for each season with reduced time spline df.

```{r}
seasons_to_test <- c("Spring", "Summer", "Fall", "Winter")
models_season <- list()
preds_season  <- list()

for (ssn in seasons_to_test) {
  season_data <- citywide %>% filter(season == ssn)
  season_data <- season_data %>%
    mutate(time_sub = as.numeric(date - min(date)))

  cb_p_season <- crossbasis(
    season_data$ppt_mean,
    lag    = 14,
    argvar = list(fun = "ns", df = 3),
    arglag = list(fun = "ns", df = 4)
  )

  cb_t_season <- crossbasis(
    season_data$temp_mean,
    lag    = 14,
    argvar = list(fun = "ns", df = 3),
    arglag = list(fun = "ns", df = 4)
  )

  season_time_df <- max(3, round(2 * n_years))

  mod_ssn <- glm(
    asthma_total ~ cb_p_season + cb_t_season +
      ns(time_sub, df = season_time_df) +
      dow,
    family = quasipoisson(link = "log"),
    data   = season_data
  )

  pred_ssn <- crosspred(
    cb_p_season, mod_ssn,
    at  = seq(0, quantile(season_data$ppt_mean, 0.99, na.rm = TRUE), length.out = 50),
    cen = 0
  )

  models_season[[ssn]] <- mod_ssn
  preds_season[[ssn]]  <- pred_ssn
}

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
season_colors <- c("Winter" = "dodgerblue", "Spring" = "forestgreen",
                   "Summer" = "darkorange", "Fall" = "brown")

for (ssn in seasons_to_test) {
  plot(preds_season[[ssn]], "overall",
       xlab = "Precipitation (mm)",
       ylab = "Cumulative RR (lag 0–14)",
       main = ssn,
       col = season_colors[ssn],
       ci.arg = list(col = adjustcolor(season_colors[ssn], 0.2)),
       lwd = 2)
  abline(h = 1, lty = 2, col = "gray50")
}
par(mfrow = c(1, 1))
```

## PM2.5 Adjustment

Three-panel comparison of cumulative exposure-response: (1) no temperature or PM2.5 adjustment, (2) primary model with temperature only, (3) primary model plus PM2.5. The contrast between panels 1 and 2 shows the impact of temperature confounding; panels 2 and 3 contrast shows the impact of further adjustment for PM2.5 (treated here as a hypothesized mediator).

```{r}
# Panel 1: no temp, no PM2.5
model_notemp <- glm(
  asthma_total ~ cb_precip +
    ns(time, df = time_df) +
    dow + season,
  family = quasipoisson(link = "log"),
  data   = citywide
)
pred_notemp <- crosspred(
  cb_precip, model_notemp,
  at  = seq(0, max(citywide$ppt_mean, na.rm = TRUE), length.out = 100),
  cen = 0
)

# Panel 3: primary + PM2.5
model_pm25 <- glm(
  asthma_total ~ cb_precip + cb_temp +
    ns(time, df = time_df) +
    pm25_mean +
    dow + season,
  family = quasipoisson(link = "log"),
  data   = citywide
)
pred_pm25 <- crosspred(
  cb_precip, model_pm25,
  at  = seq(0, max(citywide$ppt_mean, na.rm = TRUE), length.out = 100),
  cen = 0
)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

plot(pred_notemp, "overall",
     xlab = "Precipitation (mm)",
     ylab = "Cumulative RR (lag 0–14)",
     main = "No temp, no PM2.5",
     col = "gray40",
     ci.arg = list(col = adjustcolor("gray40", 0.2)),
     lwd = 2)
abline(h = 1, lty = 2, col = "gray50")

plot(pred_primary, "overall",
     xlab = "Precipitation (mm)",
     ylab = "Cumulative RR (lag 0–14)",
     main = "Primary (with temp)",
     col = "steelblue",
     ci.arg = list(col = adjustcolor("steelblue", 0.2)),
     lwd = 2)
abline(h = 1, lty = 2, col = "gray50")

plot(pred_pm25, "overall",
     xlab = "Precipitation (mm)",
     ylab = "Cumulative RR (lag 0–14)",
     main = "Primary + PM2.5",
     col = "darkred",
     ci.arg = list(col = adjustcolor("darkred", 0.2)),
     lwd = 2)
abline(h = 1, lty = 2, col = "gray50")

par(mfrow = c(1, 1))

cat("\nPM2.5 coefficient (log):",
    round(coef(model_pm25)["pm25_mean"], 5), "\n")
cat("PM2.5 IRR:", round(exp(coef(model_pm25)["pm25_mean"]), 4), "\n")
```

# Model Diagnostics

Deviance residual plots for the primary model.

```{r}
res <- residuals(model_primary, type = "deviance")

# Match residuals to data rows (glm drops NA rows silently)
res_idx   <- as.integer(names(res))
res_dates <- citywide$date[res_idx]
res_time  <- citywide$time[res_idx]

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Residuals over time
plot(res_dates, res,
     pch = ".", col = adjustcolor("steelblue", 0.5),
     xlab = "Date", ylab = "Deviance Residuals",
     main = "Residuals Over Time")
loess_fit  <- loess(res ~ res_time, span = 0.1, na.action = na.exclude)
loess_pred <- predict(loess_fit, newdata = data.frame(res_time = res_time))
lines(res_dates, loess_pred, col = "red")
abline(h = 0, lty = 2)

# Residuals vs fitted
plot(fitted(model_primary), res,
     pch = ".", col = adjustcolor("steelblue", 0.5),
     xlab = "Fitted Values", ylab = "Deviance Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

# QQ plot
qqnorm(res, pch = ".", col = "steelblue", main = "Q-Q Plot of Residuals")
qqline(res, col = "red")

# ACF of residuals
acf(res, lag.max = 21, main = "ACF of Residuals")

par(mfrow = c(1, 1))
```
