---
title: "Borough Level Analysis: Exploring Interactions"
author: "Maya Arnott"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# create outputs folder if it doesn't already exist
if (!dir.exists("outputs")) {
  dir.create("outputs")
}
```

## 1. Data Prepration

```{r, include=FALSE}
library(stringr)
library(MASS)
library(lmtest)
library(car)
library(broom)
library(ggplot2)
library(marginaleffects)
library(AER)
library(lubridate)
library(ggplot2)
library(readr)
library(dplyr)
library(tidycensus)
library(tidyverse)
library(ICC)
library(gt)

asthma_precip_pm25 <- read_csv("datasets/asthma_precip_pm25_clean.csv")

```

#### Prepare the variables
```{r}
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    date        = as.Date(date),
    borough     = as.factor(borough),
    year        = as.factor(year(date)),
    day_of_week = factor(wday(date, label = TRUE), ordered = FALSE),
    season      = as.factor(case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall"
    ))
  )

# create log transformed ppt and pm25 var
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(
    log_ppt = log(ppt + 1)
  )

asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(
    pm25 = if_else(pm25 < 0, 0, pm25),
    log_pm25 = log(pm25 + 1)
  )

# create thresholds for ppt
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(ppt_cat = case_when(
    ppt == 0        ~ "none",
    ppt < 1         ~ "trace",
    ppt < 10        ~ "light",
    ppt < 25        ~ "moderate",
    ppt >= 25       ~ "heavy"
  ) |> factor(levels = c("none", "trace", "light", "moderate", "heavy")))

```

An offset population variable was created because the raw counts can be misleading. Data was obtained from the`tidycensus` package.

```{r}
census_api_key(Sys.getenv("CENSUS_API_KEY"), overwrite = TRUE)

# pop for years 2016 to 2019
years <- c(2016, 2017, 2018, 2019)

borough_pop <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B01003_001",  # total population estimate
    state     = "NY",
    county    = c("New York", "Kings", "Queens", "Bronx", "Richmond"),
    year      = .x
  ) |>
    mutate(year = as.character(.x))
}) |>
  mutate(borough = case_when(
    grepl("New York County", NAME)  ~ "Manhattan",
    grepl("Kings County", NAME)     ~ "Brooklyn",
    grepl("Queens County", NAME)    ~ "Queens",
    grepl("Bronx County", NAME)     ~ "Bronx",
    grepl("Richmond County", NAME)  ~ "Staten Island"
  )) |>
  select(borough, year, population = estimate)

# merge with other data
asthma_precip_pm25 <- asthma_precip_pm25 |>
  left_join(borough_pop, by = c("borough", "year"))
```

Temperature is likely a confounder in this relationship, and should be added to the model. Through NOAA, there is only data available from one weather station in Central Park, so the assumption is that the temperature across boroughs is fairly similar.

```{r}
# NOAA CDO exports tmax/tmin in degrees C
temp_daily <- read_csv("datasets/daily_temp_nyc.csv") |>
  mutate(
    date     = as.Date(DATE),
    temp_avg = (TMAX / 10 + TMIN / 10) / 2
  ) |>
  select(date, temp_avg)

asthma_precip_pm25 <- asthma_precip_pm25 |>
  left_join(temp_daily, by = "date")
```

#### Scale variables

`log_ppt` and `temperature` are scaled because this helps models converge and makes coefficients across covariates comparable. `log_pm25` is scaled here as well for use in the PM2.5-adjusted sensitivity analysis only. It is excluded from the main models because PM2.5 is a likely mediator on the causal pathway (precipitation washes particulate matter from the air, so adjusting for it in the primary model would constitute over-adjustment).

```{r}
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    log_ppt_scaled  = as.numeric(scale(log_ppt)),
    log_pm25_scaled = as.numeric(scale(log_pm25)),
    temp_scaled     = as.numeric(scale(temp_avg))
  )
```

## 2. Negative Binomial Model over Poisson
```{r}
# poisson
pois.model <- glm(
  asthma_er_count ~ log_ppt_scaled + temp_scaled + borough + season +
    day_of_week + year + offset(log(population)),
  family = poisson(link = "log"),
  data = asthma_precip_pm25
)

dispersiontest(pois.model)
```

According to the overdispersion test, there is evidence of significant overdispersion (dispersion = 2.20, p<.001), which justifies the use of a negative binomial model.

## 3. ICC

```{r}
# how much do daily visits cluster within borough?
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(asthma_rate = (asthma_er_count / population) * 100000)

asthma.ICC <- ICCest(borough, asthma_rate, data = asthma_precip_pm25)
asthma.ICC$ICC
asthma.ICC$LowerCI
asthma.ICC$UpperCI

# how much does rain cluster within borough?
ppt.ICC <- ICCest(borough, log_ppt_scaled, data = asthma_precip_pm25)
ppt.ICC$ICC
```

#### ICC Results Table
```{r}
icc_table <- data.frame(
  Variable      = c("Asthma ED Visit Rate (per 100,000)", "Log Precipitation (scaled)"),
  ICC           = c(round(asthma.ICC$ICC, 3), round(ppt.ICC$ICC, 3)),
  Lower_CI      = c(round(asthma.ICC$LowerCI, 3), NA),
  Upper_CI      = c(round(asthma.ICC$UpperCI, 3), NA),
  Interpretation = c(
    "68% of variation is between boroughs",
    "No borough-level clustering"
  )
)

icc_table |>
  gt() |>
  cols_label(
    Variable       = "Variable",
    ICC            = "ICC",
    Lower_CI       = "95% CI Lower",
    Upper_CI       = "95% CI Upper",
    Interpretation = "Interpretation"
  ) |>
  tab_header(title = "Table 1. Intraclass Correlation Coefficients by Borough")
```


Asthma Rate ICC = 0.68. 68% of the total variation in daily asthma ED visit rates is attributable to between-borough differences, whereas only 32% is day-to-day within-borough variation. This justifies including borough fixed intercepts.

Precipitation ICC = -0.001. This means that precipitation is not confounded by borough and the variation in rain is only temporal.


## 4. Between borough model (ecological)

```{r}
dta_btwn <- asthma_precip_pm25 |>
  group_by(borough) |>
  summarize(
    mean_asthma_rate = mean(asthma_rate, na.rm = TRUE),
    mean_log_ppt     = mean(log_ppt_scaled, na.rm = TRUE),
    .groups = "drop"
  )

mod.btwn <- lm(
  mean_asthma_rate ~ mean_log_ppt,
  data = dta_btwn
)

summary(mod.btwn)
```

With 2 degrees of freedom, the model has very low statistical power, which justifies the use of a fixed intercept model.

## 5. Main Model: Borough Fixed Intercepts (no interactions)

```{r}
mod.fixed <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + temp_scaled + borough +
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# model outputs
summary(mod.fixed)

beta_ppt <- summary(mod.fixed)$coefficients["log_ppt_scaled", "Estimate"]
se_ppt   <- summary(mod.fixed)$coefficients["log_ppt_scaled", "Std. Error"]
lci_ppt  <- beta_ppt - 1.96 * se_ppt
uci_ppt  <- beta_ppt + 1.96 * se_ppt

rr_ppt <- paste0(
  round(exp(beta_ppt), 3), " (95% CI: ",
  round(exp(lci_ppt), 3), ", ",
  round(exp(uci_ppt), 3), ")"
)
rr_ppt
```

#### Main Model Results Table

```{r}

mod.fixed |>
  tidy(conf.int = TRUE, exponentiate = TRUE) |>
  filter(!str_detect(term, "day_of_week")) |> 
  mutate(
    term = case_when(
      term == "(Intercept)"          ~ "Intercept",
      term == "log_ppt_scaled"       ~ "Log Precipitation (1 SD)",
      term == "temp_scaled"          ~ "Temperature (1 SD)",
      term == "boroughBrooklyn"      ~ "Borough: Brooklyn vs Bronx",
      term == "boroughManhattan"     ~ "Borough: Manhattan vs Bronx",
      term == "boroughQueens"        ~ "Borough: Queens vs Bronx",
      term == "boroughStaten Island" ~ "Borough: Staten Island vs Bronx",
      term == "seasonSpring"         ~ "Season: Spring vs Fall",
      term == "seasonSummer"         ~ "Season: Summer vs Fall",
      term == "seasonWinter"         ~ "Season: Winter vs Fall",
      term == "year2017"             ~ "Year: 2017 vs 2016",
      term == "year2018"             ~ "Year: 2018 vs 2016",
      term == "year2019"             ~ "Year: 2019 vs 2016",
      TRUE ~ term
    )
  ) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  cols_label(
    term      = "Variable",
    estimate  = "Rate Ratio",
    conf.low  = "95% CI Lower",
    conf.high = "95% CI Upper",
    p.value   = "p-value"
  ) |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 3) |>
  fmt(columns = p.value, fns = function(x) ifelse(x < 0.001, "<0.001", round(x, 3))) |>
  tab_header(title = "Table 2. Borough Fixed Intercept Negative Binomial Model") |>
  tab_footnote("Rate ratios and 95% CIs shown. Reference categories: Bronx (borough), Fall (season), 2016 (year). Offset = log(population).")
```

#### Fixed Intercept Coefficient Plot

```{r}

coef_plot_df <- mod.fixed |>
  tidy(conf.int = TRUE, exponentiate = FALSE) |>  # keep on log scale for plot
  filter(term != "(Intercept)") |>
  filter(!str_detect(term, "day_of_week")) |> 
  mutate(term = case_when(
    term == "log_ppt_scaled"            ~ "Log Precipitation (1 SD)",
    term == "temp_scaled"               ~ "Temperature (1 SD)",
    term == "boroughBrooklyn"           ~ "Borough: Brooklyn vs Bronx",
    term == "boroughManhattan"          ~ "Borough: Manhattan vs Bronx",
    term == "boroughQueens"             ~ "Borough: Queens vs Bronx",
    term == "boroughStaten Island"      ~ "Borough: Staten Island vs Bronx",
    term == "seasonSpring"              ~ "Season: Spring vs Fall",
    term == "seasonSummer"              ~ "Season: Summer vs Fall",
    term == "seasonWinter"              ~ "Season: Winter vs Fall",
    term == "year2017"                  ~ "Year: 2017 vs 2016",
    term == "year2018"                  ~ "Year: 2018 vs 2016",
    term == "year2019"                  ~ "Year: 2019 vs 2016",
    TRUE ~ term
  ))

ggplot(coef_plot_df, aes(x = estimate, y = reorder(term, estimate))) + 
  geom_vline(
    xintercept = 0,
    linetype   = "dashed",
    color      = "grey55",
    linewidth  = 0.6
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    width     = 0.25,
    linewidth = 0.8,
    color     = "#3B528B"
  ) +
  geom_point(color = "#3B528B", size = 3) +
  labs(
    title    = "Fixed Intercept Negative Binomial Model Coefficients",
    subtitle = "± 95% confidence interval",
    x        = "Log Rate Ratio (effect on asthma ED visit rate)",
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12, margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 9, color = "#555555", lineheight = 1.3, margin = margin(b = 8)),
    panel.grid.minor = element_blank()
  ) 

# save plot as png
ggsave(
  filename = "outputs/coef_plot_fixed_intercept.png",
  plot     = last_plot(),
  width    = 8,
  height   = 6,
  dpi      = 300
)
```

**`log_ppt_scaled`** A 1 SD increase in log precipitation is associated with a 0.7% decrease in asthma ED visit rates (RR = 0.993, 95% CI: 0.988–0.998, p = 0.005). The direction is negative — rain may reduce same-day asthma burden by washing PM2.5 and ozone from the air. The effect is statistically significant but small in magnitude.

**`temp_scaled`** A 1 SD increase in daily temperature is associated with a 1.3% increase in asthma ED visit rates (RR = 1.013, 95% CI: 1.005–1.021, p = 0.002). Because season is already in the model, this captures within-season day-to-day thermal variation — warmer days relative to the seasonal average carry modestly higher asthma risk, likely through heat-related airway inflammation and higher ozone exposure.

**`borough`** All boroughs had significantly lower asthma ED visit rates than the Bronx (all p < 0.001). Rate ratios: Brooklyn 0.68, Manhattan 0.54, Queens 0.30, Staten Island 0.57. The Bronx's dominant baseline is consistent with documented asthma burden disparities driven by housing quality, green space, and pollution exposure.

**`season`** Spring was slightly elevated relative to Fall (RR = 1.03, p < 0.001), consistent with allergy season overlap. Summer was protective (RR = 0.77, p < 0.001) even after adjusting for temperature, suggesting seasonal factors beyond heat. Winter was modestly lower than Fall (RR = 0.97, p = 0.002).

#### Marginal Effects plot
```{r}
# scale parameters for back-transformation to mm
log_ppt_mean <- mean(asthma_precip_pm25$log_ppt, na.rm = TRUE)
log_ppt_sd   <- sd(asthma_precip_pm25$log_ppt, na.rm = TRUE)

# clip at 99th percentile
ppt_cap <- quantile(asthma_precip_pm25$ppt, 0.99, na.rm = TRUE)
ppt_seq <- seq(
  min(asthma_precip_pm25$log_ppt_scaled, na.rm = TRUE),
  (log(ppt_cap + 1) - log_ppt_mean) / log_ppt_sd,
  length.out = 200
)

# compute predictions directly from model coefficients + vcov (delta method)
# population offset cancels when converting count to rate
coefs <- coef(mod.fixed)
vc    <- vcov(mod.fixed)

borough_terms <- c(
  "Bronx"         = NA,
  "Brooklyn"      = "boroughBrooklyn",
  "Manhattan"     = "boroughManhattan",
  "Queens"        = "boroughQueens",
  "Staten Island" = "boroughStaten Island"
)

marg_preds <- purrr::map_dfr(names(borough_terms), function(bor) {
  bor_term <- borough_terms[bor]
  if (is.na(bor_term)) {
    terms <- c("(Intercept)", "log_ppt_scaled")
    W     <- cbind(1, ppt_seq)
  } else {
    terms <- c("(Intercept)", bor_term, "log_ppt_scaled")
    W     <- cbind(1, 1, ppt_seq)
  }
  lp    <- as.numeric(W %*% coefs[terms])
  se_lp <- sqrt(diag(W %*% vc[terms, terms] %*% t(W)))
  data.frame(
    borough  = bor,
    ppt_mm   = exp(ppt_seq * log_ppt_sd + log_ppt_mean) - 1,
    rate     = 100000 * exp(lp),
    rate_low = 100000 * exp(lp - 1.96 * se_lp),
    rate_hi  = 100000 * exp(lp + 1.96 * se_lp)
  )
})

# observed precipitation for tick marks on plot
ppt_obs <- asthma_precip_pm25 |>
  distinct(date, ppt) |>
  filter(ppt <= ppt_cap)

borough_colors <- c(
  "Bronx"         = "#E63946",
  "Brooklyn"      = "#F4A261",
  "Manhattan"     = "#2A9D8F",
  "Queens"        = "#3B528B",
  "Staten Island" = "#8B5E3C"
)

ggplot(marg_preds, aes(x = ppt_mm, y = rate, color = borough, fill = borough)) +
  geom_ribbon(aes(ymin = rate_low, ymax = rate_hi), alpha = 0.12, color = NA) +
  geom_line(linewidth = 1) +
  geom_rug(
    data        = ppt_obs,
    aes(x = ppt),
    inherit.aes = FALSE,
    color       = "grey50",
    alpha       = 0.3,
    linewidth   = 0.3,
    sides       = "b"
  ) +
  scale_color_manual(values = borough_colors) +
  scale_fill_manual(values = borough_colors) +
  coord_cartesian(xlim = c(0, ppt_cap)) +
  labs(
    title    = "Predicted Asthma ED Visit Rate by Precipitation and Borough",
    subtitle = "Parallel slopes indicate no borough × precipitation interaction; shading = 95% CI; tick marks = observed daily precipitation; x-axis capped at 99th percentile",
    x        = "Daily Precipitation (mm)",
    y        = "Predicted Asthma ED Visits per 100,000",
    color    = "Borough",
    fill     = "Borough"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12, margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 9, color = "#555555", lineheight = 1.3, margin = margin(b = 8)),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )

ggsave(
  filename = "outputs/marginal_effects_ppt_by_borough.png",
  plot     = last_plot(),
  width    = 9,
  height   = 5,
  dpi      = 300
)
```


## 6. Effect Modification - Borough Interaction 

```{r}

# reduced model (as noted earlier)
mod.fixed <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + temp_scaled + borough +
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# full model with interaction
mod.full <- glm.nb(
  asthma_er_count ~ log_ppt_scaled * borough + temp_scaled +
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# likelihood ratio test
lrtest(mod.fixed, mod.full)

# anova test
Anova(mod.full, type = "III")
```

Given the large borough coefficients and highly significant z values, an interaction model was compared to the main model. The LRT p-value is 0.6492, which means that the interaction term does not significantly improve the model fit. The borough disparities appear to be driven by chronic structural differences, not differing sensitivity to varying levels of precipitation. 


## 7. Sensitivity Analyses

### 7a. PM2.5-Adjusted Sensitivity Analysis

PM2.5 is excluded from the primary models because it is a likely mediator (precipitation washes particulate matter from the air, so adjusting for it would block part of the causal mechanism and produce over-adjustment bias). This sensitivity analysis adds `log_pm25_scaled` to assess how much the precipitation estimate changes.

```{r}
# PM2.5-adjusted fixed intercept model
mod.pm25 <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + log_pm25_scaled + temp_scaled + borough +
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

summary(mod.pm25)
```

The precipitation estimate is essentially stable: RR = 0.992 (β = −0.008) compared to RR = 0.993 (β = −0.007) in the primary model. This is a negligible shift that confirms PM2.5 is neither confounding nor meaningfully mediating the precipitation-asthma relationship. The decision to exclude PM2.5 from the primary model is validated.

PM2.5 itself is not significantly associated with asthma ED visits after adjusting for season and temperature (β = −0.003, p = 0.244). The non-significant and counter-directional coefficient suggests that season and temperature already absorb most of the day-to-day PM2.5 variation, leaving little residual signal. Note that 26 observations were dropped due to PM2.5 missingness in this model.

### 7b. Categorical Precipitation Sensitivity Analysis

```{r}

# categorical ppt model
model.cat <- glm.nb(
  asthma_er_count ~ ppt_cat + temp_scaled + borough + season +
    day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# AIC
summary(model.cat)
```

***Category Threshold Model***

Only light and moderate rain are significant, whereas heavy rain was not significant.

#### Sensitivity Analysis Summary Table

```{r}
# helper to extract RR and 95% CI string for a given term
extract_rr <- function(model, term) {
  coefs <- summary(model)$coefficients
  if (!term %in% rownames(coefs)) return(list(rr = NA, ci = NA, p = NA))
  b  <- coefs[term, "Estimate"]
  se <- coefs[term, "Std. Error"]
  list(
    rr = round(exp(b), 3),
    ci = paste0(round(exp(b - 1.96 * se), 3), "–", round(exp(b + 1.96 * se), 3)),
    p  = round(coefs[term, "Pr(>|z|)"], 3)
  )
}

ppt_fixed <- extract_rr(mod.fixed, "log_ppt_scaled")
ppt_full  <- extract_rr(mod.full,  "log_ppt_scaled")
ppt_pm25  <- extract_rr(mod.pm25,  "log_ppt_scaled")

# categorical model: one row per non-reference level
cat_coefs <- summary(model.cat)$coefficients
cat_terms <- rownames(cat_coefs)[grepl("ppt_cat", rownames(cat_coefs))]
cat_rows <- purrr::map_dfr(cat_terms, function(term) {
  b  <- cat_coefs[term, "Estimate"]
  se <- cat_coefs[term, "Std. Error"]
  label <- gsub("ppt_cat", "", term)
  data.frame(
    Model       = paste0("  ", label),
    RR          = round(exp(b), 3),
    CI          = paste0(round(exp(b - 1.96 * se), 3), "–", round(exp(b + 1.96 * se), 3)),
    p           = round(cat_coefs[term, "Pr(>|z|)"], 3),
    Notes       = "",
    stringsAsFactors = FALSE
  )
})

sens_table <- bind_rows(
  data.frame(
    Model = "1. NB Fixed Intercepts (primary)",
    RR    = ppt_fixed$rr,
    CI    = ppt_fixed$ci,
    p     = ppt_fixed$p,
    Notes = "Primary model",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Model = "2. NB + Borough × Precipitation Interaction",
    RR    = ppt_full$rr,
    CI    = ppt_full$ci,
    p     = ppt_full$p,
    Notes = "LRT vs. primary: p = 0.649",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Model = "3. NB + PM2.5 Adjusted",
    RR    = ppt_pm25$rr,
    CI    = ppt_pm25$ci,
    p     = ppt_pm25$p,
    Notes = "PM2.5 not significant (p = 0.244)",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Model = "4. NB Categorical Precipitation (vs. no rain)",
    RR    = NA, CI = NA, p = NA,
    Notes = "Tests threshold effects",
    stringsAsFactors = FALSE
  ),
  cat_rows
)

sens_table |>
  gt() |>
  cols_label(
    Model = "Model / Precipitation Term",
    RR    = "Rate Ratio",
    CI    = "95% CI",
    p     = "p-value",
    Notes = "Notes"
  ) |>
  fmt_number(columns = RR, decimals = 3) |>
  fmt(columns = p, fns = function(x) ifelse(is.na(x), "—", ifelse(x < 0.001, "<0.001", sprintf("%.3f", x)))) |>
  fmt(columns = c(RR, CI), fns = function(x) ifelse(is.na(x), "—", as.character(x))) |>
  tab_style(
    style     = cell_fill(color = "lightblue"),
    locations = cells_body(rows = 1)
  ) |>
  tab_style(
    style     = cell_text(indent = px(15)),
    locations = cells_body(rows = Model %in% cat_rows$Model)
  ) |>
  tab_header(title = "Table 3. Precipitation Effect Across Models") |>
  tab_footnote("Rate ratios for log_ppt_scaled (1 SD increase in log precipitation) except Model 4 which shows category-specific RRs vs. no rain. All models adjust for temperature, borough, season, day of week, and year with log(population) offset.")
```

#### Time series plot of asthma rates by borough

```{r}
# time series plot
ggplot(asthma_precip_pm25, aes(x = date, y = asthma_rate, color = borough)) +
  geom_smooth(
    method  = "loess",
    span    = 0.15,
    se      = FALSE,
    linewidth = 0.9
  ) +
  geom_line(alpha = 0.15, linewidth = 0.3) +
  scale_color_manual(values = c(
    "Bronx"         = "#E63946",
    "Brooklyn"      = "#F4A261",
    "Manhattan"     = "#2A9D8F",
    "Queens"        = "#3B528B",
    "Staten Island" = "#8B5E3C"
  )) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  facet_wrap(~ borough, ncol = 1, scales = "free_y") +
  labs(
    title    = "Daily Asthma ED Visit Rates by Borough, 2016–2019",
    x        = NULL,
    y        = "Asthma ED Visits per 100,000",
    color    = "Borough"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13, margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 9, color = "#555555", margin = margin(b = 8)),
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    strip.text       = element_text(face = "bold", size = 10)
  )

# save
ggsave(
  filename = "outputs/timeseries_asthma_storm_events.png",
  plot     = last_plot(),
  width    = 10,
  height   = 12,
  dpi      = 300
)
```


