---
title: "Maya's Model Building"
author: "Maya Arnott"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Data Prepration

#### Libraries and load in our data
```{r}
library(stringr)
library(MASS)
library(lmtest)
library(car)
library(ggplot2)
library(marginaleffects)
library(AER)
library(lubridate)
library(ggplot2)
library(readr)
library(dplyr)
library(tidycensus)
library(tidyverse)
library(ICC)

asthma_precip_pm25 <- read_csv("datasets/asthma_precip_pm25_clean.csv")

```

#### Prepare the variables
```{r}
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    date        = as.Date(date),
    borough     = as.factor(borough),
    year        = as.factor(year(date)),
    day_of_week = as.factor(wday(date, label = TRUE)),
    season      = as.factor(case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall"
    ))
  )

# create log transformed ppt var
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(log_ppt = log(ppt + 1))

# create thresholds for ppt
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(ppt_cat = case_when(
    ppt == 0        ~ "none",
    ppt < 1         ~ "trace",
    ppt < 10        ~ "light",
    ppt < 25        ~ "moderate",
    ppt >= 25       ~ "heavy"
  ) |> factor(levels = c("none", "trace", "light", "moderate", "heavy")))

```

I created an offset population variable because the raw counts can be misleading. I used data from the`tidycensus` package to do this.

```{r}
census_api_key("4469a6e8d98ef67d16af4d4b943bebffce79071e", install = TRUE)

# pop for years 2016 to 2019
years <- c(2016, 2017, 2018, 2019)

borough_pop <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B01003_001",  # total population estimate
    state     = "NY",
    county    = c("New York", "Kings", "Queens", "Bronx", "Richmond"),
    year      = .x
  ) |>
    mutate(year = as.character(.x))
}) |>
  mutate(borough = case_when(
    grepl("New York County", NAME)  ~ "Manhattan",
    grepl("Kings County", NAME)     ~ "Brooklyn",
    grepl("Queens County", NAME)    ~ "Queens",
    grepl("Bronx County", NAME)     ~ "Bronx",
    grepl("Richmond County", NAME)  ~ "Staten Island"
  )) |>
  select(borough, year, population = estimate)

# merge with other data
asthma_precip_pm25 <- asthma_precip_pm25 |>
  left_join(borough_pop, by = c("borough", "year"))
```

Temperature is likely a confounder in this relationship, and should be added to the model. I am assuming that the temperature across boroughs is fairly similar, and therefore only using data from one weather station in Central Park for average temperature will suffice.

```{r}
# will add and merge daily average temperature data to csv file

```

#### Scale variables

I scaled `log_ppt`, `pm25`, and `temperature` because this helps mixed models converge and makes coefficients across covariates comparable.
```{r}
# scale variables 
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    log_ppt_scaled = as.numeric(scale(log_ppt)),
    pm25_scaled    = as.numeric(scale(pm25)),
    # add temperature to scale
    )
```

## 2. Negative Binomial Model over Poisson
```{r}
# poisson
pois.model <- glm(
  asthma_er_count ~ log_ppt_scaled + pm25_scaled + borough + season +
    day_of_week + year + offset(log(population)),
  family = poisson(link = "log"),
  data = asthma_precip_pm25
)

dispersiontest(pois.model)
```

According to the overdispersion test, there is evidence of significant overdispersion (dispersion = 2.19, p<.001), which justifies the use of a negative binomial model.

##3. ICC

```{r}
# how much do daily visits cluster within borough?
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(asthma_rate = (asthma_er_count / population) * 100000)

asthma.ICC <- ICCest(borough, asthma_rate, data = asthma_precip_pm25)
asthma.ICC$ICC
asthma.ICC$LowerCI
asthma.ICC$UpperCI

# how much does rain cluster within borough?
ppt.ICC <- ICCest(borough, log_ppt_scaled, data = asthma_precip_pm25)
ppt.ICC$ICC
```


Asthma Rate ICC = 0.68. 68% of the total variation in daily asthma ED visit rates is attributable to between-borough differences, whereas only 32% is day-to-day within-borough variation. This justifies including borough fixed intercepts.

Precipitation ICC = -0.001. This means that precipitation is not confounded by borough and the variation in rain is only temporal.

##4. Between borough model (ecological)

```{r}
dta_btwn <- asthma_precip_pm25 |> 
  group_by(borough) |> 
  summarize(
    mean_asthma_rate = mean(asthma_rate, na.rm = TRUE),
    mean_log_ppt     = mean(log_ppt_scaled, na.rm = TRUE),
    mean_pm25        = mean(pm25_scaled, na.rm = TRUE),
    .groups = "drop"
  )

mod.btwn <- lm(
  mean_asthma_rate ~ mean_log_ppt + mean_pm25,
  data = dta_btwn
)

summary(mod.btwn)
```

With 2 degrees of freedom, the model has very low statistical power, which justifies the use of a fixed intercept model.

##5. Main Model: Borough Fixed Intercepts (no interactions)

```{r}
mod.fixed <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + pm25_scaled + borough + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# model outputs
summary(mod.fixed)

beta_ppt <- summary(mod.fixed)$coefficients["log_ppt_scaled", "Estimate"]
se_ppt   <- summary(mod.fixed)$coefficients["log_ppt_scaled", "Std. Error"]
lci_ppt  <- beta_ppt - 1.96 * se_ppt
uci_ppt  <- beta_ppt + 1.96 * se_ppt

rr_ppt <- paste0(
  round(exp(beta_ppt), 3), " (95% CI: ",
  round(exp(lci_ppt), 3), ", ",
  round(exp(uci_ppt), 3), ")"
)
rr_ppt
```

**`log_ppt_scaled`** A 1 standard deviation increase in log precipitation is associated with a 0.8% decrease in asthma ED visit rates. This is statistically significant but indicates a very small effect on asthma ED visit rates.

**`pm25_scaled`** The p-value was not significant.

**Borough fixed-intercepts** All boroughs had an association with asthma ED visits that was were highly significant. Bronx had the highest asthma ED visit rate, and all other boroughs were significantly lower.

**Season** All seasons had an association with asthma ED visits that was were highly significant. The summer season was highly protective and the spring season showed a slightly elevated association with asthma ED visits, possibly due to an overlap with allergy season.

##6. Effect Modification - Borough Interaction 

```{r}

# reduced model (as noted earlier)
mod.fixed <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + pm25_scaled + borough + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# full model with interaction
mod.full <- glm.nb(
  asthma_er_count ~ log_ppt_scaled * borough + pm25_scaled + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# likelihood ratio test
lrtest(mod.fixed, mod.full)

# anova test
Anova(model.full, type = "III")
```

The LRT p-value is 0.6492, which means that the interaction term does not significantly improve the model fit. Although we proved that boroughs have different baseline asthma rates, precipitation affects them similarly and the association does not differ significantly across boroughs.

## 7. Categorical Precipitation Sensitivity Analysis

```{r}

mod.full <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + borough + pm25_scaled + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# categorical ppt model
model.cat <- glm.nb(
  asthma_er_count ~ ppt_cat + borough + pm25_scaled + season + 
    day_of_week + year + offset(log(population)), 
  data = asthma_precip_pm25
)

# AIC
summary(model.cat)
```

***Category Threshold Model***

Only light and moderate rain are significant, whereas heavy rain was not significant.

Could there be a lag effect of precipitation? Could it be that mold or humidity spikes a couple days after rain? Or perhaps there is no significant association. It could be the case that on heavy rain days people stay home and avoid the ED, even if they do have symptoms.

## 8. Storm Events Analysis

I want to add storm event type data from the NOAA Storm Events database.

```{r}
noaa_storms <- read_csv("datasets/noaa_storm_data.csv") |> 
  janitor::clean_names() |> 
  filter(event_type %in% c(
    "Heavy Rain", 
    "Flash Flood", 
    "Flood", 
    "Tropical Storm", 
    "Tropical Depression",
    "Coastal Flood"
  )) |>  # standardize all borough names
   mutate(borough = case_when(
    str_detect(cz_name_str, regex("BRONX", ignore_case = TRUE)) ~ "Bronx",
    str_detect(cz_name_str, regex("KINGS|BROOKLYN", ignore_case = TRUE)) ~ "Brooklyn",
    str_detect(cz_name_str, regex("NEW YORK|MANHATTAN", ignore_case = TRUE)) ~ "Manhattan",
    str_detect(cz_name_str, regex("QUEENS", ignore_case = TRUE)) ~ "Queens",
    str_detect(cz_name_str, regex("RICHMOND|STATEN", ignore_case = TRUE)) ~ "Staten Island",
    TRUE ~ NA_character_
  ))

# convert begin and end dates to date class
noaa_storms <- noaa_storms |> 
  mutate(
  begin_date = mdy(begin_date), 
  end_date = mdy(end_date)
) |> 
  # expand each storm over its date range and borough
  rowwise() |> 
  mutate(date = list(seq.Date(begin_date, end_date, by = "day"))) |> 
  unnest(date) |> 
  ungroup() |> 
  select(borough, event_type, date, everything())

```

```{r}
storm_events <- noaa_storms |> 
  select(borough, date, event_type) |> 
  group_by(borough, date) |> 
  summarize(
     storm_event = paste(unique(event_type), collapse = ", "),
    .groups = "drop"
  )

# merge with asthma_precip_pm25
asthma_model_df <- asthma_precip_pm25 |> 
  left_join(storm_events |> select(borough, date, storm_event), by = c("borough", "date")) |> 
   mutate(
    storm_event = if_else(is.na(storm_event), 0, 1)
   )

# create lags for storm events

asthma_model_df <- asthma_model_df |> 
  arrange(borough, date) |> 
  group_by(borough) |> 
  mutate(
    post_storm1 = lag(storm_event, 1, default = 0),
    post_storm2 = lag(storm_event, 2, default = 0),
    post_storm3 = lag(storm_event, 3, default = 0),
    storm_cum3  = storm_event + post_storm1 + post_storm2
  ) |> 
  ungroup()
  )
  

# convert to factors
asthma_model_df <- asthma_model_df |> 
  mutate(
    storm_event = factor(storm_event, levels = c(0,1), labels = c("None","Storm")),
    post_storm1 = factor(post_storm1, levels = c(0,1), labels = c("None","PostStorm1")),
    post_storm2 = factor(post_storm2, levels = c(0,1), labels = c("None","PostStorm2")),
    post_storm3 = factor(post_storm3, levels = c(0,1), labels = c("None","PostStorm3"))
  )


# run model with storm lag
model.storm.lag <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + pm25_scaled +
    storm_event + post_storm1 + post_storm2 +
    borough + season + day_of_week + year +
    offset(log(population)),
  data = asthma_model_df
)

summary(model.storm.lag)
```

