---
title: "Fixed Intercept and Interaction Models"
author: "Maya Arnott"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# create outputs folder if it doesn't already exist
if (!dir.exists("outputs")) {
  dir.create("outputs")
}
```

## 1. Data Prepration

#### Libraries and load in our data
```{r, include=FALSE}
library(stringr)
library(MASS)
library(lmtest)
library(car)
library(broom)
library(ggplot2)
library(marginaleffects)
library(AER)
library(lubridate)
library(ggplot2)
library(readr)
library(dplyr)
library(tidycensus)
library(tidyverse)
library(ICC)
library(gt)

asthma_precip_pm25 <- read_csv("datasets/asthma_precip_pm25_clean.csv")

```

#### Prepare the variables
```{r}
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    date        = as.Date(date),
    borough     = as.factor(borough),
    year        = as.factor(year(date)),
    day_of_week = as.factor(wday(date, label = TRUE)),
    season      = as.factor(case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall"
    ))
  )

# create log transformed ppt and pm25 var
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(
    log_ppt = log(ppt + 1)
  )

asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(
    pm25 = if_else(pm25 < 0, 0, pm25),
    log_pm25 = log(pm25 + 1)
  )

# create thresholds for ppt
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(ppt_cat = case_when(
    ppt == 0        ~ "none",
    ppt < 1         ~ "trace",
    ppt < 10        ~ "light",
    ppt < 25        ~ "moderate",
    ppt >= 25       ~ "heavy"
  ) |> factor(levels = c("none", "trace", "light", "moderate", "heavy")))

```

I created an offset population variable because the raw counts can be misleading. I used data from the`tidycensus` package to do this.

```{r}
census_api_key(Sys.getenv("CENSUS_API_KEY"), overwrite = TRUE)

# pop for years 2016 to 2019
years <- c(2016, 2017, 2018, 2019)

borough_pop <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B01003_001",  # total population estimate
    state     = "NY",
    county    = c("New York", "Kings", "Queens", "Bronx", "Richmond"),
    year      = .x
  ) |>
    mutate(year = as.character(.x))
}) |>
  mutate(borough = case_when(
    grepl("New York County", NAME)  ~ "Manhattan",
    grepl("Kings County", NAME)     ~ "Brooklyn",
    grepl("Queens County", NAME)    ~ "Queens",
    grepl("Bronx County", NAME)     ~ "Bronx",
    grepl("Richmond County", NAME)  ~ "Staten Island"
  )) |>
  select(borough, year, population = estimate)

# merge with other data
asthma_precip_pm25 <- asthma_precip_pm25 |>
  left_join(borough_pop, by = c("borough", "year"))
```

Temperature is likely a confounder in this relationship, and should be added to the model. I am assuming that the temperature across boroughs is fairly similar, and therefore only using data from one weather station in Central Park for average temperature will suffice.

```{r}
# will add and merge daily average temperature data to csv file

```

#### Scale variables

I scaled `log_ppt`, `log_pm25`, and `temperature` because this helps mixed models converge and makes coefficients across covariates comparable.

```{r}
# scale variables 
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    log_ppt_scaled = as.numeric(scale(log_ppt)),
    log_pm25_scaled    = as.numeric(scale(log_pm25))
    # add temperature to scale
    )

```

## 2. Negative Binomial Model over Poisson
```{r}
# poisson
pois.model <- glm(
  asthma_er_count ~ log_ppt_scaled + log_pm25_scaled + borough + season +
    day_of_week + year + offset(log(population)),
  family = poisson(link = "log"),
  data = asthma_precip_pm25
)

dispersiontest(pois.model)
```

According to the overdispersion test, there is evidence of significant overdispersion (dispersion = 2.20, p<.001), which justifies the use of a negative binomial model.

## 3. ICC

```{r}
# how much do daily visits cluster within borough?
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(asthma_rate = (asthma_er_count / population) * 100000)

asthma.ICC <- ICCest(borough, asthma_rate, data = asthma_precip_pm25)
asthma.ICC$ICC
asthma.ICC$LowerCI
asthma.ICC$UpperCI

# how much does rain cluster within borough?
ppt.ICC <- ICCest(borough, log_ppt_scaled, data = asthma_precip_pm25)
ppt.ICC$ICC
```

#### ICC Results Table
```{r}
icc_table <- data.frame(
  Variable      = c("Asthma ED Visit Rate (per 100,000)", "Log Precipitation (scaled)"),
  ICC           = c(round(asthma.ICC$ICC, 3), round(ppt.ICC$ICC, 3)),
  Lower_CI      = c(round(asthma.ICC$LowerCI, 3), NA),
  Upper_CI      = c(round(asthma.ICC$UpperCI, 3), NA),
  Interpretation = c(
    "68% of variation is between boroughs",
    "No borough-level clustering"
  )
)

icc_table |>
  gt() |>
  cols_label(
    Variable       = "Variable",
    ICC            = "ICC",
    Lower_CI       = "95% CI Lower",
    Upper_CI       = "95% CI Upper",
    Interpretation = "Interpretation"
  ) |>
  tab_header(title = "Table 1. Intraclass Correlation Coefficients by Borough")
```


Asthma Rate ICC = 0.68. 68% of the total variation in daily asthma ED visit rates is attributable to between-borough differences, whereas only 32% is day-to-day within-borough variation. This justifies including borough fixed intercepts.

Precipitation ICC = -0.001. This means that precipitation is not confounded by borough and the variation in rain is only temporal.


## 4. Between borough model (ecological)

```{r}
dta_btwn <- asthma_precip_pm25 |> 
  group_by(borough) |> 
  summarize(
    mean_asthma_rate = mean(asthma_rate, na.rm = TRUE),
    mean_log_ppt     = mean(log_ppt_scaled, na.rm = TRUE),
    mean_pm25        = mean(log_pm25_scaled, na.rm = TRUE),
    .groups = "drop"
  )

mod.btwn <- lm(
  mean_asthma_rate ~ mean_log_ppt + mean_pm25,
  data = dta_btwn
)

summary(mod.btwn)
```

With 2 degrees of freedom, the model has very low statistical power, which justifies the use of a fixed intercept model.

## 5. Main Model: Borough Fixed Intercepts (no interactions)

```{r}
mod.fixed <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + log_pm25_scaled + borough + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# model outputs
summary(mod.fixed)

beta_ppt <- summary(mod.fixed)$coefficients["log_ppt_scaled", "Estimate"]
se_ppt   <- summary(mod.fixed)$coefficients["log_ppt_scaled", "Std. Error"]
lci_ppt  <- beta_ppt - 1.96 * se_ppt
uci_ppt  <- beta_ppt + 1.96 * se_ppt

rr_ppt <- paste0(
  round(exp(beta_ppt), 3), " (95% CI: ",
  round(exp(lci_ppt), 3), ", ",
  round(exp(uci_ppt), 3), ")"
)
rr_ppt
```

#### Main Model Results Table

```{r}

mod.fixed |>
  tidy(conf.int = TRUE, exponentiate = TRUE) |>
  filter(!str_detect(term, "day_of_week")) |> 
  mutate(
    term = case_when(
      term == "(Intercept)"          ~ "Intercept",
      term == "log_ppt_scaled"       ~ "Log Precipitation (1 SD)",
      term == "log_pm25_scaled"      ~ "Log PM2.5 (1 SD)",
      term == "boroughBrooklyn"      ~ "Borough: Brooklyn vs Bronx",
      term == "boroughManhattan"     ~ "Borough: Manhattan vs Bronx",
      term == "boroughQueens"        ~ "Borough: Queens vs Bronx",
      term == "boroughStaten Island" ~ "Borough: Staten Island vs Bronx",
      term == "seasonSpring"         ~ "Season: Spring vs Fall",
      term == "seasonSummer"         ~ "Season: Summer vs Fall",
      term == "seasonWinter"         ~ "Season: Winter vs Fall",
      term == "year2017"             ~ "Year: 2017 vs 2016",
      term == "year2018"             ~ "Year: 2018 vs 2016",
      term == "year2019"             ~ "Year: 2019 vs 2016",
      TRUE ~ term
    )
  ) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  cols_label(
    term      = "Variable",
    estimate  = "Rate Ratio",
    conf.low  = "95% CI Lower",
    conf.high = "95% CI Upper",
    p.value   = "p-value"
  ) |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 3) |>
  fmt(columns = p.value, fns = function(x) ifelse(x < 0.001, "<0.001", round(x, 3))) |>
  tab_header(title = "Table 2. Borough Fixed Intercept Negative Binomial Model") |>
  tab_footnote("Rate ratios and 95% CIs shown. Reference categories: Bronx (borough), Fall (season), 2016 (year). Offset = log(population).")
```

#### Fixed Intercept Coefficient Plot

```{r}

coef_plot_df <- mod.fixed |>
  tidy(conf.int = TRUE, exponentiate = FALSE) |>  # keep on log scale for plot
  filter(term != "(Intercept)") |>
  filter(!str_detect(term, "day_of_week")) |> 
  mutate(term = case_when(
    term == "log_ppt_scaled"            ~ "Log Precipitation (1 SD)",
    term == "log_pm25_scaled"           ~ "Log PM2.5 (1 SD)",
    term == "boroughBrooklyn"           ~ "Borough: Brooklyn vs Bronx",
    term == "boroughManhattan"          ~ "Borough: Manhattan vs Bronx",
    term == "boroughQueens"             ~ "Borough: Queens vs Bronx",
    term == "boroughStaten Island"      ~ "Borough: Staten Island vs Bronx",
    term == "seasonSpring"              ~ "Season: Spring vs Fall",
    term == "seasonSummer"              ~ "Season: Summer vs Fall",
    term == "seasonWinter"              ~ "Season: Winter vs Fall",
    term == "year2017"                  ~ "Year: 2017 vs 2016",
    term == "year2018"                  ~ "Year: 2018 vs 2016",
    term == "year2019"                  ~ "Year: 2019 vs 2016",
    TRUE ~ term
  ))

ggplot(coef_plot_df, aes(x = estimate, y = reorder(term, estimate))) + 
  geom_vline(
    xintercept = 0,
    linetype   = "dashed",
    color      = "grey55",
    linewidth  = 0.6
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    width     = 0.25,
    linewidth = 0.8,
    color     = "#3B528B"
  ) +
  geom_point(color = "#3B528B", size = 3) +
  labs(
    title    = "Fixed Intercept Negative Binomial Model Coefficients",
    subtitle = "± 95% confidence interval",
    x        = "Log Rate Ratio (effect on asthma ED visit rate)",
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12, margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 9, color = "#555555", lineheight = 1.3, margin = margin(b = 8)),
    panel.grid.minor = element_blank()
  ) 

# save plot as png
ggsave(
  filename = "outputs/coef_plot_fixed_intercept.png",
  plot     = last_plot(),
  width    = 8,
  height   = 6,
  dpi      = 300
)
```

**`log_ppt_scaled`** A 1 standard deviation increase in log precipitation is associated with a 0.75% decrease in asthma ED visit rates, holding all other variables constant. This is statistically significant but indicates a very small effect on asthma ED visit rates.

**`log_pm25_scaled`** The p-value was not significant.

**'borough'** All boroughs had an association with asthma ED visits that was were highly significant. Bronx had the highest asthma ED visit rate, and all other boroughs were significantly lower.

**'season'** All seasons had an association with asthma ED visits that was were highly significant. The summer season was highly protective and the spring season showed a slightly elevated association with asthma ED visits, possibly due to an overlap with allergy season.

#### Marginal Effects plot
```{r}
# include a marginal effects plot for mod.fixed
```


## 6. Effect Modification - Borough Interaction 

```{r}

# reduced model (as noted earlier)
mod.fixed <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + log_pm25_scaled + borough + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# full model with interaction
mod.full <- glm.nb(
  asthma_er_count ~ log_ppt_scaled * borough + log_pm25_scaled + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# likelihood ratio test
lrtest(mod.fixed, mod.full)

# anova test
Anova(mod.full, type = "III")
```

The LRT p-value is 0.6492, which means that the interaction term does not significantly improve the model fit. Although we proved that boroughs have different baseline asthma rates, precipitation affects them similarly and the association does not differ significantly across boroughs.


## 7. Categorical Precipitation Sensitivity Analysis

```{r}

mod.full <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + borough + log_pm25_scaled + 
    season + day_of_week + year + offset(log(population)),
  data = asthma_precip_pm25
)

# categorical ppt model
model.cat <- glm.nb(
  asthma_er_count ~ ppt_cat + borough + log_pm25_scaled + season + 
    day_of_week + year + offset(log(population)), 
  data = asthma_precip_pm25
)

# AIC
summary(model.cat)
```

***Category Threshold Model***

Only light and moderate rain are significant, whereas heavy rain was not significant.

## 8. Storm Events Analysis

I am adding storm event date and type data from the NOAA Storm Events database.

```{r}
noaa_storms <- read_csv("datasets/noaa_storm_data.csv") |> 
  janitor::clean_names() |> 
  filter(event_type %in% c(
    "Heavy Rain", 
    "Flash Flood", 
    "Flood", 
    "Tropical Storm", 
    "Tropical Depression",
    "Coastal Flood"
  )) |>  # standardize all borough names
   mutate(borough = case_when(
    str_detect(cz_name_str, regex("BRONX", ignore_case = TRUE)) ~ "Bronx",
    str_detect(cz_name_str, regex("KINGS|BROOKLYN", ignore_case = TRUE)) ~ "Brooklyn",
    str_detect(cz_name_str, regex("NEW YORK|MANHATTAN", ignore_case = TRUE)) ~ "Manhattan",
    str_detect(cz_name_str, regex("QUEENS", ignore_case = TRUE)) ~ "Queens",
    str_detect(cz_name_str, regex("RICHMOND|STATEN", ignore_case = TRUE)) ~ "Staten Island",
    TRUE ~ NA_character_
  ))

# convert begin and end dates to date class
noaa_storms <- noaa_storms |> 
  mutate(
  begin_date = mdy(begin_date), 
  end_date = mdy(end_date)
) |> 
  # expand each storm over its date range and borough
  rowwise() |> 
  mutate(date = list(seq.Date(begin_date, end_date, by = "day"))) |> 
  unnest(date) |> 
  ungroup() |> 
  select(borough, event_type, date, everything())

```

#### Merge and create lag after storm type.
```{r}
storm_events <- noaa_storms |> 
  select(borough, date, event_type) |> 
  group_by(borough, date) |> 
  summarize(
     storm_event = paste(unique(event_type), collapse = ", "),
    .groups = "drop"
  )

# merge with asthma_precip_pm25
asthma_model_df <- asthma_precip_pm25 |> 
  left_join(storm_events |> select(borough, date, storm_event), by = c("borough", "date")) |> 
   mutate(
    storm_event = if_else(is.na(storm_event), 0, 1)
   )

# create lags for storm events
asthma_model_df <- asthma_model_df |> 
  arrange(borough, date) |> 
  group_by(borough) |> 
  mutate(
    post_storm1 = lag(storm_event, 1, default = 0),
    post_storm2 = lag(storm_event, 2, default = 0),
    post_storm3 = lag(storm_event, 3, default = 0),
    storm_cum3  = storm_event + post_storm1 + post_storm2
  ) |> 
  ungroup()

# convert to factors
asthma_model_df <- asthma_model_df |> 
  mutate(
    storm_event = factor(storm_event, levels = c(0,1), labels = c("None","Storm")),
    post_storm1 = factor(post_storm1, levels = c(0,1), labels = c("None","PostStorm1")),
    post_storm2 = factor(post_storm2, levels = c(0,1), labels = c("None","PostStorm2")),
    post_storm3 = factor(post_storm3, levels = c(0,1), labels = c("None","PostStorm3"))
  )

# run model with storm lag
model.storm.lag <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + log_pm25_scaled +
    storm_event + post_storm1 + post_storm2 +
    borough + season + day_of_week + year +
    offset(log(population)),
  data = asthma_model_df
)

summary(model.storm.lag)
```

On storm days, asthma ER visits are significantly lower. There is a 6.6% decrease in asthma ER visits on storm days. The lagged effects a day after the storm (`post_storm1`) are not significant (p = 0.525), and we see the same story two days after a storm (`post_storm2`, p = 0.589).

#### Quantifying cumulative exposure over 3 days post-storm.

```{r}
mod.storm.cumulative <- glm.nb(
  asthma_er_count ~ log_ppt_scaled + log_pm25_scaled + storm_cum3 +
    borough + season + day_of_week + year +
    offset(log(population)),
  data = asthma_model_df
)

summary(mod.storm.cumulative)
```

#### Model Comparison Table
``` {r}
model_comparison <- data.frame(
  Model = c(
    "1. Poisson (baseline)",
    "2. NB Fixed Intercepts (primary)",
    "3. NB + Borough Interaction",
    "4. NB Categorical Precipitation",
    "5. NB + Storm Events"
  ),
  AIC = c(
    AIC(pois.model),
    AIC(mod.fixed),
    AIC(mod.full),
    AIC(model.cat),
    AIC(model.storm.lag)
  ),
  Df = c(
    pois.model$df.residual,
    mod.fixed$df.residual,
    mod.full$df.residual,
    model.cat$df.residual,
    model.storm.lag$df.residual
  ),
  Notes = c(
    "Overdispersion confirmed, NB preferred",
    "Primary model",
    "LRT p=0.62, interaction not significant",
    "Sensitivity analysis",
    "Storm day effect: -6.6%, lags not significant"
  )
)

model_comparison |>
  gt() |>
  cols_label(
    Model = "Model",
    AIC   = "AIC",
    Df    = "Residual df",
    Notes = "Notes"
  ) |>
  fmt_number(columns = AIC, decimals = 1) |>
  tab_header(title = "Table 3. Model Comparison Summary") |>
  tab_style(
    style = cell_fill(color = "lightblue"),
    locations = cells_body(rows = 2)  # highlight primary model
  )
```

#### Time series plot of asthma rates with storm events indicated

```{r}
# get unique storm event dates
storm_dates <- asthma_model_df |>
  filter(storm_event == "Storm") |>
  pull(date) |>
  unique()

# time series plot
ggplot(asthma_model_df, aes(x = date, y = asthma_rate, color = borough)) +
  # vertical lines = storm events
  geom_vline(
    xintercept = storm_dates,
    color      = "darkblue",
    linewidth  = 0.3,
    alpha      = 0.5
  ) +
  geom_smooth(
    method  = "loess",
    span    = 0.15,
    se      = FALSE,
    linewidth = 0.9
  ) +
  geom_line(alpha = 0.15, linewidth = 0.3) +
  scale_color_manual(values = c(
    "Bronx"         = "#E63946",
    "Brooklyn"      = "#F4A261",
    "Manhattan"     = "#2A9D8F",
    "Queens"        = "#3B528B",
    "Staten Island" = "#8B5E3C"
  )) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  facet_wrap(~ borough, ncol = 1, scales = "free_y") +
  labs(
    title    = "Daily Asthma ED Visit Rates by Borough, 2016–2019",
    subtitle = "Dark blue vertical lines indicate storm event days (heavy rain, flood, tropical storm, or coastal flood)",
    x        = NULL,
    y        = "Asthma ED Visits per 100,000",
    color    = "Borough"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13, margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 9, color = "#555555", margin = margin(b = 8)),
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    strip.text       = element_text(face = "bold", size = 10)
  )

# save
ggsave(
  filename = "outputs/timeseries_asthma_storm_events.png",
  plot     = last_plot(),
  width    = 10,
  height   = 12,
  dpi      = 300
)
```

