---
title: "Maya's Model Building"
author: "Maya Arnott"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries and load in our data
```{r}
library(stringr)
library(MASS)
library(lmtest)
library(car)
library(ggplot2)
library(marginaleffects)
library(AER)
library(lubridate)
library(ggplot2)
library(readr)
library(dplyr)
library(tidycensus)

asthma_precip_pm25 <- read_csv("datasets/asthma_precip_pm25_clean.csv")

```

Prepare the variables
```{r}
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(
    date        = as.Date(date),
    borough     = as.factor(borough),
    year        = as.factor(year(date)),
    day_of_week = as.factor(wday(date, label = TRUE)),
    season      = as.factor(case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall"
    ))
  )

# create log transformed ppt var
asthma_precip_pm25 <- asthma_precip_pm25 |> 
  mutate(log_ppt = log(ppt + 1))

# create thresholds for ppt
asthma_precip_pm25 <- asthma_precip_pm25 |>
  mutate(ppt_cat = case_when(
    ppt == 0        ~ "none",
    ppt < 1         ~ "trace",
    ppt < 10        ~ "light",
    ppt < 25        ~ "moderate",
    ppt >= 25       ~ "heavy"
  ) |> factor(levels = c("none", "trace", "light", "moderate", "heavy")))

```

I created an offset population variable because the raw counts can be misleading. I used data from the`tidycensus` package to do this.

```{r}
census_api_key("4469a6e8d98ef67d16af4d4b943bebffce79071e", install = TRUE)

# pop for years 2016 to 2019
years <- c(2016, 2017, 2018, 2019)

borough_pop <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B01003_001",  # total population estimate
    state     = "NY",
    county    = c("New York", "Kings", "Queens", "Bronx", "Richmond"),
    year      = .x
  ) |>
    mutate(year = as.character(.x))
}) |>
  mutate(borough = case_when(
    grepl("New York County", NAME)  ~ "Manhattan",
    grepl("Kings County", NAME)     ~ "Brooklyn",
    grepl("Queens County", NAME)    ~ "Queens",
    grepl("Bronx County", NAME)     ~ "Bronx",
    grepl("Richmond County", NAME)  ~ "Staten Island"
  )) |>
  select(borough, year, population = estimate)

# merge with other data
asthma_precip_pm25 <- asthma_precip_pm25 |>
  left_join(borough_pop, by = c("borough", "year"))
```


Visualize relationship again
```{r}
ggplot(asthma_precip_pm25, 
       aes(x = ppt, y = asthma_er_count)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") +
  facet_wrap(~ borough) +
  theme_minimal() +
  labs(title = "Precipitation vs Asthma ED Visits by Borough")
```

Simple baseline model(s)
```{r}
# continuous log-transformed ppt model
model.log <- glm.nb(
  asthma_er_count ~ log_ppt + borough + season + day_of_week + year + pm25 + offset(log(population)) ,
  data = asthma_precip_pm25
)

# categorical ppt model
model.cat <- glm.nb(
  asthma_er_count ~ ppt_cat + borough + season + day_of_week + year + pm25 + offset(log(population)), 
  data = asthma_precip_pm25
)

# compare AIC and summaries
AIC(model.log, model.cat)

summary(model.log)
summary(model.cat)

```

The two models have very similar AIC's (53191 for the log transformed model vs 53193 for the categorical threshold model).

***Log Model***  

Interestingly, the `log_ppt` coefficient has a slight negative p-value of -.007, with a significant p-value, meaning that more rain is weakly associated with fewer ED visits. Precipitation alone is a weak predictor of asthma ED visits compared to borough, season, and day of week. Each doubling of precipitation is associated with a 0.5% decrease in the asthma ED visit rate.

***Category Threshold Model***

Only light and moderate rain are significant and are also negative, whereas heavy rain was not significant. 

***Across Models***
Borough effects are huge and highly significant across the models, with Queens having the lowest rate and reference category, the Bronx, having the highest. I'd like to formally test the interaction borough has on the relationship between asthma and precipitation. 


Could there be a lag effect of precipitation? Could it be that mold or humidity spikes a couple days after rain? Or perhaps there is no significant association. It could be the case that on heavy rain days people stay home and avoid the ED, even if they do have symptoms.

Interaction model with `borough`

```{r}
# reduced model (as noted earlier)
model.log <- glm.nb(
  asthma_er_count ~ log_ppt + borough + season + day_of_week + year +  offset(log(population)),
  data = asthma_precip_pm25
)

# full model with interaction
model.full <- glm.nb(
  asthma_er_count ~ log_ppt * borough + season + day_of_week + year + offset(log(population)), 
  data = asthma_precip_pm25
)

# likelihood ratio test
lrtest(model.log, model.full)

# anova test
Anova(model.full, type = "III")
```

The LRT p-value is 0.62, which means that the interaction term does not significantly improve the model fit and suggests precipitation affects asthma ED visits similarly across all five boroughs. The reduced model with no interaction is preferred.

Exploring a season interaction model

```{r}
# is ppt effect modified by season? 
model.season.int <- glm.nb(
  asthma_er_count ~ log_ppt * season + borough + day_of_week + year +
    offset(log(population)),
  data = asthma_precip_pm25
)

lrtest(model.log, model.season.int)
```

I want to add storm event type data from the NOAA Storm Events database.

```{r}
noaa_storms <- read_csv("datasets/noaa_storm_data.csv") |> 
  janitor::clean_names() |> 
  filter(event_type %in% c(
    "Heavy Rain", 
    "Flash Flood", 
    "Flood", 
    "Tropical Storm", 
    "Tropical Depression",
    "Coastal Flood"
  )) |>  # standardize all borough names
   mutate(borough = case_when(
    str_detect(cz_name_str, regex("BRONX", ignore_case = TRUE)) ~ "Bronx",
    str_detect(cz_name_str, regex("KINGS|BROOKLYN", ignore_case = TRUE)) ~ "Brooklyn",
    str_detect(cz_name_str, regex("NEW YORK|MANHATTAN", ignore_case = TRUE)) ~ "Manhattan",
    str_detect(cz_name_str, regex("QUEENS", ignore_case = TRUE)) ~ "Queens",
    str_detect(cz_name_str, regex("RICHMOND|STATEN", ignore_case = TRUE)) ~ "Staten Island",
    TRUE ~ NA_character_
  ))

# convert begin and end dates to date class
noaa_storms <- noaa_storms |> 
  mutate(
  begin_date = mdy(begin_date), 
  end_date = mdy(end_date)
) |> 
  # expand each storm over its date range and borough
  rowwise() |> 
  mutate(date = list(seq.Date(begin_date, end_date, by = "day"))) |> 
  unnest(date) |> 
  ungroup() |> 
  select(borough, event_type, date, everything())

```

```{r}
storm_events <- noaa_storms |> 
  select(borough, date, event_type) |> 
  group_by(borough, date) |> 
  summarize(
     storm_event = paste(unique(event_type), collapse = ", "),
    .groups = "drop"
  )

# merge with asthma_precip_pm25
asthma_model_df <- asthma_precip_pm25 |> 
  left_join(storm_events, by = c("borough", "date")) |> 
  # replace NA storm_event with "None"
  mutate(storm_event = ifelse(is.na(storm_event), "None", storm_event))

# create lag for storm events (add more lags here later)

asthma_model_df <- asthma_model_df |> 
  mutate(
    # create a binary storm event for that day
    storm_event = if_else(storm_event != "None", 1, 0)
  ) |> 
  group_by(borough) |> 
  arrange(date) |>    
  mutate(
    post_storm1 = lag(storm_event, 1)  # 1-day lag
  ) |> 
  ungroup() |> 
  # replace NA lag with 0
  mutate(post_storm1 = ifelse(is.na(post_storm1), 0, post_storm1))

# create storm indicators and lags then convert to factors

asthma_model_df <- asthma_model_df |> 
  # create storm indicators and lags
  mutate(
    storm_event = if_else(date %in% storm_event, 1, 0),
    post_storm1 = lag(storm_event, 1, default = 0),
    post_storm2 = lag(storm_event, 2, default = 0),
    post_storm3 = lag(storm_event, 3, default = 0),
    # cumulative storm effect over 3 days
    storm_cum3 = storm_event + post_storm1 + post_storm2
  ) |> 
  mutate(
    storm_event = factor(storm_event, levels = c(0,1), labels = c("None","Storm")),
    post_storm1 = factor(post_storm1, levels = c(0,1), labels = c("None","PostStorm1")),
    post_storm2 = factor(post_storm2, levels = c(0,1), labels = c("None","PostStorm2")),
    post_storm3 = factor(post_storm3, levels = c(0,1), labels = c("None","PostStorm3"))
  )

# run model with storm lag
model.storm.lag <- glm.nb(
  asthma_er_count ~ log_ppt + storm_event + post_storm1 + post_storm2 + borough + season + day_of_week + year + offset(log(population)),
  data = asthma_model_df
)

```

