---
title: "Project Data Visualization"
output: html_document
date: "2026-03-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(here)
library(readr)
library(splines)
library(mgcv)
```

## Importing cleaned + merged data
```{r data}
data <- read.csv("asthma_precip_clean.csv")
```

## Categorizing dates by seasons, years, months, day of the week

```{r time variables, echo=TRUE}
##Convert Date Variable
data$date <- as.Date(data$date)

#Create season variable
data <- data %>%
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    )
  )

#Create Year var, Month var, and Day of Week var.
data <- data %>%
  mutate(
    date = as.Date(date),

    # Year
    year = year(date),

    # Month (numeric and labeled)
    month_num = month(date),
    month = month(date, label = TRUE, abbr = FALSE),

    # Day of week (ordered Monday–Sunday)
    day_of_week = wday(date, label = TRUE, abbr = FALSE),
    day_of_week = factor(day_of_week,
                         levels = c("Monday", "Tuesday", "Wednesday",
                                    "Thursday", "Friday",
                                    "Saturday", "Sunday"))
  )
```

## Summary statistics (mean, sd, median, min, max, quartiles) of Asthma ER admissions and PPT by borough and time variables

```{r summary statistics, echo=FALSE}
#Citywide and all time data
data %>%
  summarise(
    # ASTHMA
    mean_asthma   = mean(asthma_er_count, na.rm = TRUE),
    sd_asthma     = sd(asthma_er_count, na.rm = TRUE),
    min_asthma    = min(asthma_er_count, na.rm = TRUE),
    Q1_asthma     = quantile(asthma_er_count, 0.25, na.rm = TRUE),
    median_asthma = median(asthma_er_count, na.rm = TRUE),
    Q3_asthma     = quantile(asthma_er_count, 0.75, na.rm = TRUE),
    max_asthma    = max(asthma_er_count, na.rm = TRUE),
    IQR_asthma    = IQR(asthma_er_count, na.rm = TRUE),

    # PRECIPITATION
    mean_ppt   = mean(ppt, na.rm = TRUE),
    sd_ppt     = sd(ppt, na.rm = TRUE),
    min_ppt    = min(ppt, na.rm = TRUE),
    Q1_ppt     = quantile(ppt, 0.25, na.rm = TRUE),
    median_ppt = median(ppt, na.rm = TRUE),
    Q3_ppt     = quantile(ppt, 0.75, na.rm = TRUE),
    max_ppt    = max(ppt, na.rm = TRUE),
    IQR_ppt    = IQR(ppt, na.rm = TRUE)
  )

#By borough
data %>%
  group_by(borough) %>%
  summarise(
    # ASTHMA
    mean_asthma   = mean(asthma_er_count, na.rm = TRUE),
    sd_asthma     = sd(asthma_er_count, na.rm = TRUE),
    min_asthma    = min(asthma_er_count, na.rm = TRUE),
    Q1_asthma     = quantile(asthma_er_count, 0.25, na.rm = TRUE),
    median_asthma = median(asthma_er_count, na.rm = TRUE),
    Q3_asthma     = quantile(asthma_er_count, 0.75, na.rm = TRUE),
    max_asthma    = max(asthma_er_count, na.rm = TRUE),
    IQR_asthma    = IQR(asthma_er_count, na.rm = TRUE),

    # PRECIPITATION
    mean_ppt   = mean(ppt, na.rm = TRUE),
    sd_ppt     = sd(ppt, na.rm = TRUE),
    min_ppt    = min(ppt, na.rm = TRUE),
    Q1_ppt     = quantile(ppt, 0.25, na.rm = TRUE),
    median_ppt = median(ppt, na.rm = TRUE),
    Q3_ppt     = quantile(ppt, 0.75, na.rm = TRUE),
    max_ppt    = max(ppt, na.rm = TRUE),
    IQR_ppt    = IQR(ppt, na.rm = TRUE)
  )

#By Season
data %>%
  group_by(season) %>%
  summarise(
    # ASTHMA
    mean_asthma   = mean(asthma_er_count, na.rm = TRUE),
    sd_asthma     = sd(asthma_er_count, na.rm = TRUE),
    min_asthma    = min(asthma_er_count, na.rm = TRUE),
    Q1_asthma     = quantile(asthma_er_count, 0.25, na.rm = TRUE),
    median_asthma = median(asthma_er_count, na.rm = TRUE),
    Q3_asthma     = quantile(asthma_er_count, 0.75, na.rm = TRUE),
    max_asthma    = max(asthma_er_count, na.rm = TRUE),
    IQR_asthma    = IQR(asthma_er_count, na.rm = TRUE),

    # PRECIPITATION
    mean_ppt   = mean(ppt, na.rm = TRUE),
    sd_ppt     = sd(ppt, na.rm = TRUE),
    min_ppt    = min(ppt, na.rm = TRUE),
    Q1_ppt     = quantile(ppt, 0.25, na.rm = TRUE),
    median_ppt = median(ppt, na.rm = TRUE),
    Q3_ppt     = quantile(ppt, 0.75, na.rm = TRUE),
    max_ppt    = max(ppt, na.rm = TRUE),
    IQR_ppt    = IQR(ppt, na.rm = TRUE)
  )

#By year
data %>%
  group_by(year) %>%
  summarise(
    # ASTHMA
    mean_asthma   = mean(asthma_er_count, na.rm = TRUE),
    sd_asthma     = sd(asthma_er_count, na.rm = TRUE),
    min_asthma    = min(asthma_er_count, na.rm = TRUE),
    Q1_asthma     = quantile(asthma_er_count, 0.25, na.rm = TRUE),
    median_asthma = median(asthma_er_count, na.rm = TRUE),
    Q3_asthma     = quantile(asthma_er_count, 0.75, na.rm = TRUE),
    max_asthma    = max(asthma_er_count, na.rm = TRUE),
    IQR_asthma    = IQR(asthma_er_count, na.rm = TRUE),

    # PRECIPITATION
    mean_ppt   = mean(ppt, na.rm = TRUE),
    sd_ppt     = sd(ppt, na.rm = TRUE),
    min_ppt    = min(ppt, na.rm = TRUE),
    Q1_ppt     = quantile(ppt, 0.25, na.rm = TRUE),
    median_ppt = median(ppt, na.rm = TRUE),
    Q3_ppt     = quantile(ppt, 0.75, na.rm = TRUE),
    max_ppt    = max(ppt, na.rm = TRUE),
    IQR_ppt    = IQR(ppt, na.rm = TRUE)
  )

#By month
data %>%
  group_by(month_num) %>%
  summarise(
    # ASTHMA
    mean_asthma   = mean(asthma_er_count, na.rm = TRUE),
    sd_asthma     = sd(asthma_er_count, na.rm = TRUE),
    min_asthma    = min(asthma_er_count, na.rm = TRUE),
    Q1_asthma     = quantile(asthma_er_count, 0.25, na.rm = TRUE),
    median_asthma = median(asthma_er_count, na.rm = TRUE),
    Q3_asthma     = quantile(asthma_er_count, 0.75, na.rm = TRUE),
    max_asthma    = max(asthma_er_count, na.rm = TRUE),
    IQR_asthma    = IQR(asthma_er_count, na.rm = TRUE),

    # PRECIPITATION
    mean_ppt   = mean(ppt, na.rm = TRUE),
    sd_ppt     = sd(ppt, na.rm = TRUE),
    min_ppt    = min(ppt, na.rm = TRUE),
    Q1_ppt     = quantile(ppt, 0.25, na.rm = TRUE),
    median_ppt = median(ppt, na.rm = TRUE),
    Q3_ppt     = quantile(ppt, 0.75, na.rm = TRUE),
    max_ppt    = max(ppt, na.rm = TRUE),
    IQR_ppt    = IQR(ppt, na.rm = TRUE)
  )

#By day of the week
data %>%
  group_by(day_of_week) %>%
  summarise(
    # ASTHMA
    mean_asthma   = mean(asthma_er_count, na.rm = TRUE),
    sd_asthma     = sd(asthma_er_count, na.rm = TRUE),
    min_asthma    = min(asthma_er_count, na.rm = TRUE),
    Q1_asthma     = quantile(asthma_er_count, 0.25, na.rm = TRUE),
    median_asthma = median(asthma_er_count, na.rm = TRUE),
    Q3_asthma     = quantile(asthma_er_count, 0.75, na.rm = TRUE),
    max_asthma    = max(asthma_er_count, na.rm = TRUE),
    IQR_asthma    = IQR(asthma_er_count, na.rm = TRUE),

    # PRECIPITATION
    mean_ppt   = mean(ppt, na.rm = TRUE),
    sd_ppt     = sd(ppt, na.rm = TRUE),
    min_ppt    = min(ppt, na.rm = TRUE),
    Q1_ppt     = quantile(ppt, 0.25, na.rm = TRUE),
    median_ppt = median(ppt, na.rm = TRUE),
    Q3_ppt     = quantile(ppt, 0.75, na.rm = TRUE),
    max_ppt    = max(ppt, na.rm = TRUE),
    IQR_ppt    = IQR(ppt, na.rm = TRUE)
  )


```
## Histograms (distributions of asthma and ppt by borough and time variables)

```{r histograms, echo=FALSE}
ggplot(data, aes(x = asthma_er_count)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Daily Asthma ER Visits",
       x = "Asthma ER Count",
       y = "Frequency")

ggplot(data, aes(x = ppt)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Daily Precipitation",
       x = "Daily Precipitation",
       y = "Frequency")

#data is right skewed (let's log transform this)
data <- data %>%
  mutate(log_ppt = log(ppt + 1))
ggplot(data, aes(x = log_ppt)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Daily Precipitation",
       x = "Log Precipitation",
       y = "Frequency")
#data still skewed
```

# Boxplots of Asthma and PPT grouped by borough, seasons, years, months, day of the week (effect modifications analysis)

```{r boxplots_asthmappt, echo=FALSE}

library(ggplot2)
library(dplyr)
library(tidyr)

# Make sure variables are factors
data <- data %>%
  mutate(
    year = factor(year),
    month = factor(month, levels = 1:12),
    season = factor(season, levels = c("Winter","Spring","Summer","Fall")),
    day_of_week = factor(day_of_week, levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))
  )

# 1. Asthma ER Visits by Borough
ggplot(data, aes(x = borough, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Borough",
    x = "Borough",
    y = "Asthma ER Count"
  )

# 2. Precipitation by Borough
ggplot(data, aes(x = borough, y = ppt)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(
    title = "Daily Precipitation by Borough",
    x = "Borough",
    y = "Precipitation (inches)"
  )

# 3. Asthma ER Visits by Season
ggplot(data, aes(x = season, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Season",
    x = "Season",
    y = "Asthma ER Count"
  )

# 4. Precipitation by Season
ggplot(data, aes(x = season, y = ppt)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(
    title = "Daily Precipitation by Season",
    x = "Season",
    y = "Precipitation (inches)"
  )

# 5. Asthma ER Visits by Year
ggplot(data, aes(x = year, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Year",
    x = "Year",
    y = "Asthma ER Count"
  )

# 6. Precipitation by Year
ggplot(data, aes(x = year, y = ppt)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(
    title = "Daily Precipitation by Year",
    x = "Year",
    y = "Precipitation (inches)"
  )

# 7. Asthma ER Visits by Month
ggplot(data, aes(x = factor(month_num), y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Month",
    x = "Month",
    y = "Asthma ER Count"
  )

# 8. Precipitation by Month
ggplot(data, aes(x = factor(month_num), y = ppt)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(
    title = "Daily Precipitation by Month",
    x = "Month",
    y = "Precipitation (inches)"
  )

# 9. Asthma ER Visits by Day of Week
ggplot(data, aes(x = day_of_week, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Day of Week",
    x = "Day of Week",
    y = "Asthma ER Count"
  )

# 10. Precipitation by Day of Week
ggplot(data, aes(x = day_of_week, y = ppt)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(
    title = "Daily Precipitation by Day of Week",
    x = "Day of Week",
    y = "Precipitation (inches)"
  )

```

# Box plots of Asthma visits by time variables, stratified by borough and Box plots of Asthma visits by boroughs, stratified by seasons and year
```{r box plots, echo=FALSE}
# Asthma ER visits by seasons, stratified by Borough
ggplot(data, aes(x = season, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~borough) +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by seasons, Stratified by Borough",
    x = "Month",
    y = "Asthma ER Visits"
  )

# Asthma ER visits by year, stratified by Borough
ggplot(data, aes(x = factor(year), y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~borough) +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Year, Stratified by Borough",
    x = "Year",
    y = "Asthma ER Visits"
  )

# Asthma ER visits by Month, stratified by Borough
ggplot(data, aes(x = factor(month_num), y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~borough) +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by Month, Stratified by Borough",
    x = "Month",
    y = "Asthma ER Visits"
  )

# Asthma ER visits by day of the week, stratified by Borough
ggplot(data, aes(x = day_of_week, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~borough) +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by day of week, Stratified by Borough",
    x = "Month",
    y = "Asthma ER Visits"
  )

# Asthma ER visits by borough, stratified by seasons
ggplot(data, aes(x = borough, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~season) +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by boroughs, Stratified by season",
    x = "Season",
    y = "Asthma ER Visits"
  )

# Asthma ER visits by borough, stratified by year
ggplot(data, aes(x = borough, y = asthma_er_count)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~year) +
  theme_minimal() +
  labs(
    title = "Asthma ER Visits by year, Stratified by Borough",
    x = "Year",
    y = "Asthma ER Visits"
  )
```

# Scatterplots using log transformed data and non log transformed

```{r scatterp, echo=FALSE}
#create separate dataset for citywide data 
city_data <- data %>%
  group_by(date) %>%
  summarise(
    total_asthma = sum(asthma_er_count, na.rm = TRUE),
    mean_ppt = mean(ppt, na.rm = TRUE),
    mean_log_ppt = log(mean_ppt)
  )

#city wide scatterplot
ggplot(city_data, aes(x = mean_ppt, y = total_asthma)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  theme_minimal() +
  labs(title = "Precipitation vs Asthma ED Visits",
       x = "Mean Daily Precipitation",
       y = "Total Asthma ED Visits")

#city wide log scatterplot
ggplot(city_data, aes(x = mean_log_ppt, y = total_asthma)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  theme_minimal() +
  labs(title = "LOG Precipitation vs Asthma ED Visits",
       x = "Mean Daily LOG Precipitation",
       y = "Total Asthma ED Visits")

#scatterplot with trend lines of each borough
ggplot(data, aes(x = ppt, y = asthma_er_count, color = borough)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Precipitation vs Asthma ER Visits with Trend by Borough",
    x = "Daily Precipitation (inches)",
    y = "Asthma ER Count",
    color = "Borough"
  )

#log scatterplot with trend lines of each borough
ggplot(data, aes(x = log_ppt, y = asthma_er_count, color = borough)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal() +
  labs(
    title = "LOG Precipitation vs Asthma ER Visits with Trend by Borough",
    x = "Daily LOG Precipitation",
    y = "Asthma ER Count",
    color = "Borough"
  )

#scatterplot by seasons
ggplot(data, aes(x = ppt, y = asthma_er_count)) +
  geom_point(alpha = 0.5, color = "darkorange") +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  facet_wrap(~season) +
  theme_minimal() +
  labs(
    title = "Precipitation vs Asthma ER Visits by Season",
    x = "Daily Precipitation (inches)",
    y = "Asthma ER Count"
  )

#log scatterplot by seasons
ggplot(data, aes(x = log_ppt, y = asthma_er_count)) +
  geom_point(alpha = 0.5, color = "darkorange") +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  facet_wrap(~season) +
  theme_minimal() +
  labs(
    title = "LOG Precipitation vs Asthma ER Visits by Season",
    x = "Daily LOG Precipitation",
    y = "Asthma ER Count"
  )

#scatterplot by months 
ggplot(data, aes(x = ppt, y = asthma_er_count, color = factor(month_num))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Precipitation vs Asthma ER Visits Colored by Month",
    x = "Daily Precipitation (inches)",
    y = "Asthma ER Count",
    color = "Month"
  )

#log scatterplot by months 
ggplot(data, aes(x = log_ppt, y = asthma_er_count, color = factor(month_num))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal() +
  labs(
    title = "LOG Precipitation vs Asthma ER Visits Colored by Month",
    x = "Daily LOG Precipitation",
    y = "Asthma ER Count",
    color = "Month"
  )

#scatterplot using log transformed ppt
ggplot(data, aes(x = log_ppt, y = asthma_er_count)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  theme_minimal() +
  labs(
    title = "Log-Transformed Precipitation vs Asthma ER Visits",
    x = "Log(Daily Precipitation) (zeros = 0)",
    y = "Asthma ER Count"
  )
```

# Time series analysis 

```{r timeseries, echo=TRUE}
# Plot daily city total asthma over time
ggplot(city_data, aes(x = date, y = total_asthma)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Daily Asthma ER Visits in NYC (2016–2019)",
       x = "Date",
       y = "Total Asthma ER Visits")

# Plot daily city average precipitation over time
ggplot(city_data, aes(x = date, y = mean_ppt)) +
  geom_line(color = "blue") +
  theme_minimal() +
  labs(title = "Daily Precipitation in NYC (2016–2019)",
       x = "Date",
       y = "Mean Daily Precipitation")

#creating year_num variable in the city wide data set
city_data <- city_data %>%
  mutate(
    year = year(date),
    year_num = year - 2015,
    month = month(date)
  )

city_data <- city_data %>%
  arrange(date) %>%
  mutate(
    time = as.numeric(date - min(date))
  )

#calculating the number of years in dataset to determine degrees of freedom 
yr_num <- length(unique(year(city_data$date)))

#smoothing plot
ggplot(city_data, aes(date,total_asthma))+ 
  geom_smooth(color = "yellow", method = lm, formula = y ~ ns(x, df= 14)) +
  geom_smooth(color = "pink", method = lm, formula = y ~ ns(x, df= 4)) 


ggplot(city_data, aes(date,mean_ppt))+ 
  geom_smooth(color = "orange", method = lm, formula = y ~ ns(x, df= 16)) +
  geom_smooth(color = "purple", method = lm, formula = y ~ ns(x, df= 4)) 

#Poisson model
mod.ts.p <- glm(
  total_asthma ~
    mean_ppt +
    factor(month) +
    ns(date, df = 16),
  family = poisson(),
  data = city_data)

summary(mod.ts.p)
mean(city_data$total_asthma)
var(city_data$total_asthma)

#Quasi poisson model
mod.ts.qp <- glm(
  total_asthma ~
    mean_ppt +
    factor(month) +
    ns(date, df = 16),
  family = quasipoisson(),
  data = city_data
)
summary(mod.ts.qp)$dispersion

#Poisson is severely overdispersed

```


# PPT and Asthma Lag Exploration

```{r lags, echo=TRUE}

# Plot Autocorrelation
acf(city_data$mean_ppt, lag.max = 7)
acf(city_data$mean_ppt, lag.max = 21)
#No weekly correlation of rainfall

acf(city_data$total_asthma, lag.max = 7)
acf(city_data$total_asthma, lag.max = 21)
#Weekly trend of asthma ER visits

```

